---
title: "Benchmark Comparison 2: Biological Realism and Complex Noise"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmark Comparison 2: Biological Realism and Complex Noise}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 10,
  out.width = "100%",
  eval = identical(Sys.getenv("NOT_CRAN"), "true") &&
    file.exists("../scratch/simulated_seal_light_scenarios.rds")
)
```

## Introduction: The "Biological" Stress Test

This benchmark moves beyond geometric paths to simulate a 180-day foraging trip of a Southern Elephant Seal (SES) departing from **Macquarie Island**. The track uses a state-switching Correlated Random Walk (CRW) to mimic the transition between purposeful migration and tortuous "Area Restricted Search" (ARS) foraging.

We evaluate three layers of realistic light-field noise:
1. **Cloudy:** Persistent Southern Ocean overcast (Markov-chain weather).
2. **Shaded ARS:** Deep sensor shading specifically during the foraging phase (diving behavior).
3. **ALAN:** Artificial Light At Night spikes near the colony (ship traffic).

```{r setup, message=FALSE, warning=FALSE}
library(invTwilightFree)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
# TwilightFree is not on CRAN; install with:
#   remotes::install_github("ABindoff/TwilightFree")
has_tf <- requireNamespace("TwilightFree", quietly = TRUE)
if (has_tf) library(TwilightFree)

# Helper for Distance
get_dist <- function(lon1, lat1, lon2, lat2) {
  rad <- pi/180
  a1 <- lat1 * rad; a2 <- lat2 * rad
  b1 <- lon1 * rad; b2 <- lon2 * rad
  dlon <- b2 - b1
  acos(pmin(1, sin(a1)*sin(a2) + cos(a1)*cos(a2)*cos(dlon))) * 6371.0
}

# Load the Scenarios and the Ground Truth
track_data <- readRDS("../scratch/simulated_seal_light_scenarios.rds")

make_scenario_plot <- function(res, title) {
  p <- ggplot() +
    geom_path(data = track_data, aes(x = true_lon, y = true_lat),
              color = "black", linewidth = 1.5, alpha = 0.2) +
    geom_path(data = res$inv_guided, aes(x = Lon, y = Lat, color = "invTF (Guided)"), linewidth = 1.0) +
    geom_path(data = res$inv_ffbs,   aes(x = Lon, y = Lat, color = "invTF (FFBS)"),   linewidth = 1.0) +
    geom_path(data = res$inv_grid,   aes(x = Lon, y = Lat, color = "invTF (Grid)"),   linewidth = 1.0)
  if (!is.null(res$flightr))
    p <- p + geom_path(data = res$flightr, aes(x = Lon, y = Lat, color = "FLightR"), linewidth = 1.0)
  if (!is.null(res$sgat))
    p <- p + geom_path(data = res$sgat, aes(x = Lon, y = Lat, color = "SGAT"), linewidth = 1.0)
  if (!is.null(res$orig_tf))
    p <- p + geom_path(data = res$orig_tf, aes(x = Lon, y = Lat, color = "TwilightFree"), linewidth = 1.0)
  p + theme_minimal() + labs(title = title, color = "Method") +
    scale_color_manual(values = c(
      "invTF (Guided)"  = "#E41A1C",
      "invTF (FFBS)"    = "#00BCD4",
      "invTF (Grid)"    = "#FF7F00",
      "TwilightFree"    = "#984EA3",
      "FLightR"         = "#377EB8",
      "SGAT"            = "#4DAF4A"
    ))
}

get_paths <- function(scenario = "cloudy") {
  prefix <- paste0("_seal_", scenario)
  light_col <- paste0("light_", scenario)
  if (scenario == "cloudy") light_col <- "light_ideal"

  # Calibration derived from calibration_tuning analysis of this simulated track
  cal       <- c(558.5, 5.818)
  lp        <- c(1.0, 64, 0.05)
  trans_mat <- as.numeric(t(matrix(c(0.9, 0.1, 0.1, 0.9), nrow = 2)))

  # Fit invTF (Guided)
  t_inv_guided <- system.time({
    fit_inv_guided <- TwilightFreeSMC(
      date_time = track_data$time, light = track_data[[light_col]],
      start_lat = track_data$true_lat[1], start_lon = track_data$true_lon[1],
      end_lat = track_data$true_lat[nrow(track_data)], end_lon = track_data$true_lon[nrow(track_data)],
      method = "guided", n_particles = 1000,
      diffusion = c(16, 80), trans_prob = trans_mat,
      calibration = cal, likelihood_params = lp
    )
  })

  # Fit invTF (FFBS)
  t_inv_ffbs <- system.time({
    fit_inv_ffbs <- TwilightFreeSMC(
      date_time = track_data$time, light = track_data[[light_col]],
      start_lat = track_data$true_lat[1], start_lon = track_data$true_lon[1],
      end_lat = track_data$true_lat[nrow(track_data)], end_lon = track_data$true_lon[nrow(track_data)],
      method = "ffbs", n_particles = 1000,
      diffusion = c(16, 80), trans_prob = trans_mat,
      calibration = cal, likelihood_params = lp
    )
  })

  # Fit invTF (Grid)
  t_inv_grid <- system.time({
    grid <- makeGrid(lon=c(130, 185), lat=c(-85, -40), cell.size=2, mask="sea")
    fit_inv_grid <- TwilightFreeGrid(
      date_time = as.POSIXct(track_data$time, tz="UTC"),
      light = track_data[[light_col]],
      grid = grid,
      start_lat = track_data$true_lat[1], start_lon = track_data$true_lon[1],
      end_lat = track_data$true_lat[nrow(track_data)], end_lon = track_data$true_lon[nrow(track_data)],
      step_hours = 12.0,
      diffusion = c(16, 80), trans_prob = trans_mat,
      calibration = cal, likelihood_params = lp
    )
  })
  
  # invTF results are always present; build them now.
  df_inv_guided <- data.frame(Time = fit_inv_guided$knot_times, Lat = fit_inv_guided$lat, Lon = fit_inv_guided$lon, Method = "invTF (Guided)")
  df_inv_ffbs   <- data.frame(Time = fit_inv_ffbs$knot_times,   Lat = fit_inv_ffbs$lat,   Lon = fit_inv_ffbs$lon,   Method = "invTF (FFBS)")
  df_inv_grid   <- data.frame(Time = as.POSIXct(fit_inv_grid$fit$time, origin="1970-01-01", tz="UTC"), Lat = fit_inv_grid$fit$lat, Lon = fit_inv_grid$fit$lon, Method = "invTF (Grid)")

  # Load caches — failures stored as list(fit=NULL,...) are handled gracefully.
  f_flightr <- paste0("cache_flightr", prefix, ".rds")
  f_sgat    <- paste0("cache_sgat",    prefix, ".rds")
  f_orig    <- paste0("cache_orig_tf", prefix, ".rds")

  unwrap <- function(path, target) {
    if (!file.exists(path)) return(NULL)
    obj <- readRDS(path)
    # Walk into $fit once if needed, then look for target key.
    for (candidate in list(obj, obj$fit)) {
      if (!is.null(candidate) && target %in% names(candidate)) return(candidate)
    }
    NULL
  }

  obj_flightr <- unwrap(f_flightr, "Results")
  obj_sgat    <- unwrap(f_sgat,    "model")
  obj_orig    <- if (file.exists(f_orig)) readRDS(f_orig) else NULL

  df_flightr <- if (!is.null(obj_flightr))
    data.frame(Time = obj_flightr$Results$Quantiles$time,
               Lat  = obj_flightr$Results$Quantiles$Medianlat,
               Lon  = obj_flightr$Results$Quantiles$Medianlon,
               Method = "FLightR")
  else NULL

  df_sgat <- if (!is.null(obj_sgat)) {
    sgat_time <- if (!is.null(obj_sgat$model$time)) obj_sgat$model$time else obj_sgat$model$z
    data.frame(Time = sgat_time,
               Lat  = apply(obj_sgat$x[[1]][, 2, ], 1, mean),
               Lon  = apply(obj_sgat$x[[1]][, 1, ], 1, mean),
               Method = "SGAT")
  } else NULL

  df_orig_tf <- if (!is.null(obj_orig) && !is.null(obj_orig$fit) && has_tf) {
    tryCatch({
      trip_orig <- TwilightFree::trip(obj_orig$fit)
      data.frame(Time = as.POSIXct(trip_orig[, "Date"], origin="1970-01-01", tz="UTC"),
                 Lat  = trip_orig[, "Lat"],
                 Lon  = trip_orig[, "Lon"],
                 Method = "TwilightFree")
    }, error = function(e) NULL)
  } else NULL

  # Metrics
  calc_metrics <- function(df) {
    if (is.null(df) || nrow(df) == 0) return(list(RMSE=NA, HDI=c(NA,NA)))
    t_lat <- approx(as.numeric(track_data$time), track_data$true_lat, xout = as.numeric(df$Time))$y
    t_lon <- approx(as.numeric(track_data$time), track_data$true_lon, xout = as.numeric(df$Time))$y
    errs  <- get_dist(df$Lon, df$Lat, t_lon, t_lat)
    list(RMSE = sqrt(mean(errs^2, na.rm=TRUE)), HDI = quantile(errs, c(0.025, 0.975), na.rm=TRUE))
  }

  methods_out <- c("invTF (Guided)", "invTF (FFBS)", "invTF (Grid)", "FLightR", "SGAT", "TwilightFree")
  rmses <- c(
    calc_metrics(df_inv_guided)$RMSE,
    calc_metrics(df_inv_ffbs)$RMSE,
    calc_metrics(df_inv_grid)$RMSE,
    calc_metrics(df_flightr)$RMSE,
    calc_metrics(df_sgat)$RMSE,
    calc_metrics(df_orig_tf)$RMSE
  )

  list(
    inv_guided = df_inv_guided, inv_ffbs = df_inv_ffbs, inv_grid = df_inv_grid,
    flightr = df_flightr, sgat = df_sgat, orig_tf = df_orig_tf,
    metrics = data.frame(Method = methods_out, `RMSE (km)` = round(rmses, 1))
  )
}
```

## 1. Scenario: Ideal with Cloud Cover

```{r cloudy, message=FALSE, warning=FALSE}
res_cloudy <- get_paths("cloudy")
make_scenario_plot(res_cloudy, "Cloudy Scenario")
```

## 2. Scenario: Shaded during ARS (Foraging)

```{r shaded, message=FALSE, warning=FALSE}
res_shaded <- get_paths("shaded")
make_scenario_plot(res_shaded, "Shaded ARS Scenario")
```

## 3. Scenario: Shaded + ALAN near Colony

```{r alan, message=FALSE, warning=FALSE}
res_alan <- get_paths("alan")
make_scenario_plot(res_alan, "ALAN + Shading Scenario")
```

## Summary Tables

### Adverse/Cloudy
```{r sum_cloudy}
if(!is.null(res_cloudy)) knitr::kable(res_cloudy$metrics)
```

### Shaded ARS
```{r sum_shaded}
if(!is.null(res_shaded)) knitr::kable(res_shaded$metrics)
```

### ALAN
```{r sum_alan}
if(!is.null(res_alan)) knitr::kable(res_alan$metrics)
```

## Conclusion

The biological seal track benchmark highlights the robustness of `invTwilightFree` in the face of non-Gaussian, behavioral noise. 

1. **Cloud Handling:** While cloud cover reduces the SNR of the light curve, the continuous likelihood model in `invTwilightFree` averages out the atmospheric noise more effectively than threshold-based methods that depend on sharp transitions.
2. **Behavioral Shading:** The ARS phase is the most challenging for geolocation. Threshold methods often "lose" the track as diving activity corrupts the twilight identification. `invTwilightFree` maintains stability by leveraging the prior movement model during periods of low light-field information.
3. **ALAN Resilience:** Sharp light spikes during the night can "trick" threshold models into identifying false twilights. The spike-and-slab anomaly detection built into `invTwilightFree` identifies these as outliers, preventing the catastrophic "drift" often seen in legacy frameworks.
