---
title: "Calibration and Hyperparameter Tuning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibration and Hyperparameter Tuning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8,
  out.width = "100%"
)
pkgload::load_all("..")
library(ggplot2)
library(dplyr)
library(geosphere)

# Distance helper
get_dist <- function(lon1, lat1, lon2, lat2) {
  distHaversine(cbind(lon1, lat1), cbind(lon2, lat2)) / 1000
}
```

## Introduction: The "Ground Truth" Baseline

To ensure our models are correctly calibrated, we start with a "Dead Easy" track:
- **Latitudes:** -40 to -35 (Moderate Southern latitudes, far from poles/equator).
- **Season:** November (Away from Equinox).
- **Noise:** Zero noise, "Ideal" light field.

Every package should be able to recover this track with < 50km RMSE if calibrated correctly.

```{r generate_easy_track}
# Generate a simple 20-day track
set.seed(42)
times <- seq(as.POSIXct("2024-11-01", tz="UTC"), by="5 min", length.out = 20 * 24 * 12)
n <- length(times)

# Hobart to Adelaide-ish
true_lat <- seq(-42.8, -35.0, length.out = n)
true_lon <- seq(147.3, 138.6, length.out = n)

# Generate ideal light data using the package's own solar model
# (This ensures we aren't testing cross-package solar math differences yet)
light_ideal <- invTwilightFree::solar_zenith(as.numeric(times), true_lon, true_lat)
# Convert zenith to a "light" value (e.g. max 64, transition at 96)
light_ideal <- pmax(0, 64 * (96 - light_ideal) / (96 - 85))
light_ideal[light_ideal > 64] <- 64

easy_track <- data.frame(
  time = times,
  true_lat = true_lat,
  true_lon = true_lon,
  light = light_ideal
)
```

## Systematic Tuning: Guided SMC

We iterate over `diffusion`, `n_particles`, and `lambda` to find the optimal configuration.

```{r tuning_loop_invtf}
tuning_results <- expand.grid(
  diffusion = c(30, 60),
  n_particles = c(1000, 5000),
  lambda = c(1.0, 5.0, 20.0)
)

tuning_results$RMSE <- sapply(1:nrow(tuning_results), function(i) {
  fit <- TwilightFreeSMC(
    date_time = easy_track$time, 
    light = easy_track$light,
    start_lat = -42.8, start_lon = 147.3,
    end_lat = -35.0, end_lon = 138.6,
    method = "guided", 
    n_particles = tuning_results$n_particles[i],
    diffusion = tuning_results$diffusion[i],
    likelihood_params = c(tuning_results$lambda[i], 64, 0.05)
  )
  
  t_lat <- approx(as.numeric(easy_track$time), easy_track$true_lat, xout = as.numeric(fit$knot_times))$y
  t_lon <- approx(as.numeric(easy_track$time), easy_track$true_lon, xout = as.numeric(fit$knot_times))$y
  errs <- get_dist(fit$lon, fit$lat, t_lon, t_lat)
  sqrt(mean(errs^2, na.rm=TRUE))
})

knitr::kable(tuning_results)
```

## Grid HMM Optimization

For the Grid HMM, we test resolution (`cell.size`) and `step_hours`.

```{r tuning_loop_grid}
grid_tuning <- expand.grid(
  cell_size = c(0.5, 1, 2),
  step_hours = c(4, 8)
)

grid_tuning$RMSE <- sapply(1:nrow(grid_tuning), function(i) {
  grid <- TwilightFree::makeGrid(lon=c(130, 155), lat=c(-50, -30), 
                                  cell.size=grid_tuning$cell_size[i], mask="sea")
  fit <- TwilightFreeGrid(
    date_time = easy_track$time, 
    light = easy_track$light,
    grid = grid,
    start_lat = -42.8, start_lon = 147.3,
    end_lat = -35.0, end_lon = 138.6,
    step_hours = grid_tuning$step_hours[i],
    diffusion = 60
  )
  
  trip_mat <- TwilightFree::trip(fit$fit)
  df <- data.frame(
    Time = as.POSIXct(trip_mat[, 1], origin="1970-01-01", tz="UTC"),
    Lon = trip_mat[, 2],
    Lat = trip_mat[, 3]
  )
  
  t_lat <- approx(as.numeric(easy_track$time), easy_track$true_lat, xout = as.numeric(df$Time))$y
  t_lon <- approx(as.numeric(easy_track$time), easy_track$true_lon, xout = as.numeric(df$Time))$y
  errs <- get_dist(df$Lon, df$Lat, t_lon, t_lat)
  sqrt(mean(errs^2, na.rm=TRUE))
})

knitr::kable(grid_tuning)
```

## Manual Calibration Test

Since we know the "Ideal" light was generated with a transition at 96 degrees and max light of 64, we can set the calibration parameters exactly.

```{r manual_cal}
# Expected = slope * (96 - zenith) = 5.818 * 96 - 5.818 * zenith
slope_val <- 64 / (96 - 85)
intercept_val <- slope_val * 96

fit_manual <- TwilightFreeSMC(
  date_time = easy_track$time, light = easy_track$light,
  start_lat = -42.8, start_lon = 147.3,
  end_lat = -35.0, end_lon = 138.6,
  method = "guided", n_particles = 5000,
  diffusion = 60,
  calibration = c(intercept_val, slope_val)
)

t_lat <- approx(as.numeric(easy_track$time), easy_track$true_lat, xout = as.numeric(fit_manual$knot_times))$y
t_lon <- approx(as.numeric(easy_track$time), easy_track$true_lon, xout = as.numeric(fit_manual$knot_times))$y
errs_m <- get_dist(fit_manual$lon, fit_manual$lat, t_lon, t_lat)
rmse_manual <- sqrt(mean(errs_m^2, na.rm=TRUE))

print(paste("Manual Calibration RMSE:", round(rmse_manual, 2), "km"))
```

## Comparative Performance (Ideal Data)

We now run `FLightR` and `SGAT` on this same track to see their baseline accuracy.

```{r comparison_easy}
# FLightR
# (Requires twilight detection)
tw <- TwilightFree::thresholdPath(easy_track$time, easy_track$light, threshold=10)
# ... abbreviated FLightR setup ...

# SGAT
# (Using simple threshold-based initialization)
# ... abbreviated SGAT setup ...
```

## Visual Audit of Best Fits

```{r plots}
# Use best parameters found above
best_guided <- tuning_results[which.min(tuning_results$RMSE), ]
best_grid <- grid_tuning[which.min(grid_tuning$RMSE), ]

fit_guided_best <- TwilightFreeSMC(
  date_time = easy_track$time, light = easy_track$light,
  start_lat = -42.8, start_lon = 147.3,
  end_lat = -35.0, end_lon = 138.6,
  method = "guided", n_particles = best_guided$n_particles,
  diffusion = best_guided$diffusion
)

grid_best_obj <- TwilightFree::makeGrid(lon=c(130, 155), lat=c(-50, -30), 
                                    cell.size=best_grid$cell_size, mask="sea")
fit_grid_best <- TwilightFreeGrid(
  date_time = easy_track$time, light = easy_track$light,
  grid = grid_best_obj,
  start_lat = -42.8, start_lon = 147.3,
  end_lat = -35.0, end_lon = 138.6,
  step_hours = best_grid$step_hours,
  diffusion = 60
)

trip_best <- TwilightFree::trip(fit_grid_best$fit)
df_grid_best <- data.frame(
  Time = as.POSIXct(trip_best[, 1], origin="1970-01-01", tz="UTC"),
  Lon = trip_best[, 2],
  Lat = trip_best[, 3]
)

df_guided_best <- data.frame(
  Time = as.POSIXct(fit_guided_best$knot_times, origin="1970-01-01", tz="UTC"),
  Lat = fit_guided_best$lat,
  Lon = fit_guided_best$lon
)

ggplot() +
  geom_path(data = easy_track, aes(x = true_lon, y = true_lat), color = "black", linewidth = 2, alpha = 0.2) +
  geom_path(data = df_guided_best, aes(x = Lon, y = Lat), color = "cyan", linewidth = 1) +
  geom_path(data = df_grid_best, aes(x = Lon, y = Lat), color = "green", linewidth = 1) +
  theme_minimal() +
  labs(title = "Best Recovered Track (Ideal Data)", 
       subtitle = paste0("Guided RMSE: ", round(min(tuning_results$RMSE), 1), 
                        "km | Grid RMSE: ", round(min(grid_tuning$RMSE), 1), "km"))
```


## Next Steps: Iterating on Parameters

If RMSE is > 50km for either method, we need to check:
1. **Zenith threshold:** Are we assuming 96 in the model but the data has a different transition point?
2. **Diffusion:** Is 50 km/day too restrictive or too loose for this straight-line track?
3. **Likelihood parameters:** Are the spike-and-slab priors too aggressive for ideal data?

