---
title: "Comprehensive Benchmark: Accuracy, Performance, and Noise Resilience"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comprehensive Benchmark: Accuracy, Performance, and Noise Resilience}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 12,
  out.width = "100%"
)
# Ensure the local version of the package is loaded
pkgload::load_all("..")
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(geosphere)

# Helper for distance calculation
get_dist <- function(lon1, lat1, lon2, lat2) {
  distHaversine(cbind(lon1, lat1), cbind(lon2, lat2)) / 1000
}

# Helper to unwrap caches
find_fit <- function(obj, target) {
  curr <- obj
  for(i in 1:10) {
    if (target %in% names(curr)) return(curr)
    if ("fit" %in% names(curr)) curr <- curr$fit else break
  }
  return(NULL)
}

# Core benchmarking function
get_bench_data <- function(scenario = "ideal", track_data) {
  # Prefix for caches
  f_prefix <- paste0("_seal_", scenario)
  light_col <- if(scenario == "ideal") "light_ideal" else if(scenario == "cloudy") "light_ideal" else paste0("light_", scenario)
  
  trans_sticky <- matrix(c(0.9, 0.1, 0.1, 0.9), nrow=2)
  trans_flat <- as.numeric(t(trans_sticky))

  # Fit invTF (Guided)
  t_inv_guided <- system.time({
    fit_inv_guided <- TwilightFreeSMC(
      date_time = track_data$time, light = track_data[[light_col]],
      start_lat = track_data$true_lat[1], start_lon = track_data$true_lon[1],
      end_lat = track_data$true_lat[nrow(track_data)], end_lon = track_data$true_lon[nrow(track_data)],
      method = "guided", n_particles = 5000,
      diffusion = c(16, 80),
      trans_prob = trans_flat,
      calibration = c(558.5, 5.818),
      likelihood_params = c(1.0, 64, 2.0, 20.0)
    )
  })
  
  # Fit invTF (Grid)
  t_inv_grid <- system.time({
    grid <- TwilightFree::makeGrid(lon=c(130, 185), lat=c(-85, -40), cell.size=1.0, mask="sea")
    fit_inv_grid <- TwilightFreeGrid(
      date_time = as.POSIXct(track_data$time, tz="UTC"), 
      light = track_data[[light_col]], 
      grid = grid,
      start_lat = track_data$true_lat[1], start_lon = track_data$true_lon[1],
      end_lat = track_data$true_lat[nrow(track_data)], end_lon = track_data$true_lon[nrow(track_data)],
      step_hours = 4.0,
      diffusion = c(16, 80),
      trans_prob = trans_flat,
      calibration = c(558.5, 5.818),
      likelihood_params = c(1.0, 64, 2.0, 20.0)
    )
  })
  
  # Load Caches
  f_flightr <- paste0("cache_flightr", f_prefix, ".rds")
  f_sgat <- paste0("cache_sgat", f_prefix, ".rds")
  
  # Try local and vignettes/
  found_f <- FALSE
  found_s <- FALSE
  if (file.exists(f_flightr)) found_f <- TRUE
  else if (file.exists(paste0("vignettes/", f_flightr))) { f_flightr <- paste0("vignettes/", f_flightr); found_f <- TRUE }
  
  if (file.exists(f_sgat)) found_s <- TRUE
  else if (file.exists(paste0("vignettes/", f_sgat))) { f_sgat <- paste0("vignettes/", f_sgat); found_s <- TRUE }
  
  if (!found_f || !found_s) {
    warning(paste("Caches missing for", scenario))
    return(NULL)
  }
  
  c_flightr <- readRDS(f_flightr)
  c_sgat <- readRDS(f_sgat)
  
  obj_flightr <- find_fit(c_flightr, "Results")
  obj_sgat <- find_fit(c_sgat, "model")
  
  # Extract Paths
  df_inv_guided <- data.frame(
    Time = as.POSIXct(fit_inv_guided$knot_times, origin="1970-01-01", tz="UTC"), 
    Lat = fit_inv_guided$lat, 
    Lon = fit_inv_guided$lon, 
    Method = "invTF (Guided)"
  )
  
  df_inv_grid <- data.frame(
    Time = as.POSIXct(fit_inv_grid$fit$time, origin="1970-01-01", tz="UTC"), 
    Lat = fit_inv_grid$fit$lat, 
    Lon = fit_inv_grid$fit$lon, 
    Method = "invTF (Grid)"
  )
  
  if (!is.null(obj_flightr)) {
    df_flightr <- data.frame(Time = obj_flightr$Results$Quantiles$time, Lat = obj_flightr$Results$Quantiles$Medianlat, Lon = obj_flightr$Results$Quantiles$Medianlon, Method = "FLightR")
  } else {
    df_flightr <- NULL
  }
  
  if (!is.null(obj_sgat)) {
    sgat_time_val <- if (!is.null(obj_sgat$model$time)) obj_sgat$model$time else obj_sgat$model$z
    df_sgat <- data.frame(Time = sgat_time_val, Lat = apply(obj_sgat$x[[1]][,2,], 1, mean), Lon = apply(obj_sgat$x[[1]][,1,], 1, mean), Method = "SGAT")
  } else {
    df_sgat <- NULL
  }
  
  # Metrics function
  calc_metrics <- function(df, name) {
    if (is.null(df) || nrow(df) == 0) return(NULL)
    t_lat <- approx(as.numeric(track_data$time), track_data$true_lat, xout = as.numeric(df$Time))$y
    t_lon <- approx(as.numeric(track_data$time), track_data$true_lon, xout = as.numeric(df$Time))$y
    
    # Ensure latitudes are within physical bounds for distance calculation
    safe_lat <- pmax(-90, pmin(90, df$Lat))
    errs <- get_dist(df$Lon, safe_lat, t_lon, t_lat)
    
    lat_err <- abs(df$Lat - t_lat)
    lon_err <- abs(df$Lon - t_lon)
    
    data.frame(
      Method = name,
      RMSE = sqrt(mean(errs^2, na.rm=TRUE)),
      Mean_Lat_Err = mean(lat_err, na.rm=TRUE),
      Mean_Lon_Err = mean(lon_err, na.rm=TRUE),
      Time_s = NA
    )
  }
  
  m_guided <- calc_metrics(df_inv_guided, "invTF (Guided)")
  m_guided$Time_s <- t_inv_guided["elapsed"]
  
  m_grid <- calc_metrics(df_inv_grid, "invTF (Grid)")
  m_grid$Time_s <- t_inv_grid["elapsed"]
  
  if (!is.null(df_flightr)) {
    m_flightr <- calc_metrics(df_flightr, "FLightR")
    m_flightr$Time_s <- c_flightr$time["elapsed"]
  } else {
    m_flightr <- data.frame(Method="FLightR", RMSE=NA, Mean_Lat_Err=NA, Mean_Lon_Err=NA, Time_s=NA)
  }
  
  if (!is.null(df_sgat)) {
    m_sgat <- calc_metrics(df_sgat, "SGAT")
    m_sgat$Time_s <- c_sgat$time["elapsed"]
  } else {
    m_sgat <- data.frame(Method="SGAT", RMSE=NA, Mean_Lat_Err=NA, Mean_Lon_Err=NA, Time_s=NA)
  }
  
  metrics <- bind_rows(m_guided, m_grid, m_flightr, m_sgat)
  
  df_states_guided <- data.frame(
    Time = as.POSIXct(fit_inv_guided$knot_times, origin="1970-01-01", tz="UTC"),
    Prob_State_2 = fit_inv_guided$prob_state[[2]],
    Method = "invTF (Guided)",
    Scenario = scenario
  )
  
  df_states_grid <- data.frame(
    Time = as.POSIXct(fit_inv_grid$fit$time, origin="1970-01-01", tz="UTC"),
    Prob_State_2 = fit_inv_grid$fit$prob_state[[2]],
    Method = "invTF (Grid)",
    Scenario = scenario
  )

  list(
    paths = bind_rows(df_inv_guided, df_inv_grid, df_flightr, df_sgat) %>% mutate(Scenario = scenario),
    metrics = metrics %>% mutate(Scenario = scenario),
    states = bind_rows(df_states_guided, df_states_grid)
  )
}
```

## Introduction

This comprehensive benchmark evaluates `invTwilightFree` against `FLightR` and `SGAT` across four scenarios using a 180-day simulated Southern Elephant Seal track.

### Scenarios
1. **Ideal:** Pristine data, no shading from cloud, no ALAN.
2. **Cloudy:** Overcast weather.
3. **Shaded ARS:** Deep sensor shading during foraging.
4. **ALAN + Shading:** Artificial Light At Night + Foraging noise.

```{r load_data, message=FALSE, warning=FALSE}
# Path adjustment for vignette rendering
track_path <- if(file.exists("scratch/simulated_seal_light_scenarios.rds")) "scratch/simulated_seal_light_scenarios.rds" else "../scratch/simulated_seal_light_scenarios.rds"
track_data <- readRDS(track_path)
scenarios <- c("ideal", "cloudy", "shaded", "alan")
all_results <- lapply(scenarios, function(s) get_bench_data(s, track_data))
all_paths <- bind_rows(lapply(all_results, function(x) x$paths))
all_metrics <- bind_rows(lapply(all_results, function(x) x$metrics))
all_states <- bind_rows(lapply(all_results, function(x) x$states))
```

## Accuracy and Performance Summary

```{r summary_table, echo=FALSE}
summary_tab <- all_metrics %>%
  select(Scenario, Method, RMSE, Time_s) %>%
  mutate(RMSE = round(RMSE, 1), Time_s = round(Time_s, 1)) %>%
  pivot_wider(names_from = Scenario, values_from = c(RMSE, Time_s))

knitr::kable(summary_tab, caption = "Accuracy (RMSE km) and Performance (Time s)")
```

## Visual Comparison

```{r plot_trajectories, echo=FALSE, fig.height=14}
plot_scenario <- function(scen) {
  scen_paths <- all_paths %>% filter(Scenario == scen)
  
  ggplot() +
    # True track as a thick, very soft gray line
    geom_path(data = track_data, aes(x = true_lon, y = true_lat), color = "gray70", linewidth = 2.5, alpha = 0.3, lineend = "round") +
    # Estimated track as a crisp, thin, distinct line
    geom_path(data = scen_paths, aes(x = Lon, y = Lat, color = Method), linewidth = 0.7, lineend = "round") +
    facet_wrap(~Method, ncol = 2) +
    # Tufte Minimalist Theme
    theme(
      text = element_text(family = "serif", size = 14),
      panel.background = element_blank(),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold", size = 14, hjust = 0),
      axis.line = element_line(color = "black", linewidth = 0.3),
      axis.ticks = element_line(color = "black", linewidth = 0.3),
      axis.title = element_text(size = 12),
      legend.position = "none",
      plot.title = element_text(face = "bold", size = 16, hjust = 0, margin = margin(b = 15))
    ) +
    scale_color_manual(values = c("invTF (Guided)" = "#1f78b4", "invTF (Grid)" = "#e31a1c", "FLightR" = "#33a02c", "SGAT" = "#ff7f00")) +
    labs(title = paste("Scenario:", toupper(scen)), x = "Longitude", y = "Latitude")
}

(plot_scenario("ideal") / plot_scenario("cloudy"))
(plot_scenario("shaded") / plot_scenario("alan"))
```

## Behavioural State Probabilities

The `invTwilightFree` methods uniquely feature integrated multi-state behavioural modelling. In these benchmarks, a 2-state "rapid switching" transition matrix was paired with diffusion parameters `c(10, 80)` to model localized ARS (foraging) versus wide-ranging transit states.

The plot below shows the inferred marginal probability of the animal being in the high-diffusion (Transit) state at any given knot:

```{r plot_states, echo=FALSE, fig.height=8}
ggplot(all_states, aes(x = Time, y = Prob_State_2, color = Method)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~Scenario, ncol = 2) +
  theme(
    text = element_text(family = "serif", size = 14),
    panel.background = element_blank(),
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 14, hjust = 0),
    axis.line = element_line(color = "black", linewidth = 0.3),
    axis.ticks = element_line(color = "black", linewidth = 0.3),
    legend.position = "bottom"
  ) +
  labs(title = "Probability of High-Diffusion State over Time", y = "P(State = 2)", x = "Time") +
  scale_color_manual(values = c("invTF (Guided)" = "#1f78b4", "invTF (Grid)" = "#e31a1c"))
```
