---
title: "Sensor Fusion: Combining Light with Environmental Constraints"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensor Fusion: Combining Light with Environmental Constraints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 8,
  fig.height = 4.5,
  out.width  = "100%"
)
```

## Overview

The `terms` argument of `TwilightFreeGrid()` (and `TwilightFreeSMC()`) accepts a list
of `location_term()` objects.  Each term adds an additive log-likelihood contribution
over the grid at every knot, combining any number of physical constraints with the
light likelihood.

This vignette fits four models on one simulated equinox track and compares their accuracy:

| Model | Key terms |
|-------|-----------|
| 1 ŌĆö Light only | none |
| 2 ŌĆö Light + hemisphere prior | breeding-range knowledge |
| 3 ŌĆö Light + bathymetry floor | 500 m max-dive constraint |
| 4 ŌĆö Light + all three | prior + bathy + SST (via `function_source`) |

A final section demonstrates `function_source()` and `custom_rule()`.

---

## Simulated track

The simulated animal is in the **Southern Ocean at lat = ŌłÆ25┬░, lon = 150┬░** (subtropical
South Pacific).  The tag was recovered near the September equinox.  We deliberately
generate the light observations from the **northern mirror** (+25┬░) to trigger the
well-known equinox hemisphere confusion: near 22 September the solar zenith at ┬▒25┬░
differs by only ~0.36┬░, so the light-only likelihood is bimodal and the MAP ends up
in the wrong hemisphere.

```{r simulate}
library(invTwilightFree)
library(raster)

true_lon <- 150
true_lat <- -25
cal <- c(64, 64 / 90)   # intercept, slope
lp  <- c(0.5, 64, 0.05) # lambda, max_light, prob_slab

times  <- seq(as.POSIXct("2024-09-22", tz = "UTC"),
              as.POSIXct("2024-09-24", tz = "UTC"),
              by = "10 min")
t_unix <- as.numeric(times)

set.seed(42)
zenith_north <- solar_zenith(t_unix,
                              rep(true_lon, length(t_unix)),
                              rep(+25,      length(t_unix)))
obs <- pmax(0, pmin(64, cal[1] - cal[2] * zenith_north + rnorm(length(t_unix), 0, 1)))
```

The grid is symmetric around the equator so the north/south degeneracy is exact within
the grid:

```{r grid}
grid <- makeGrid(lon = c(145, 155), lat = c(-35, 35), cell.size = 2.5)
```

---

## Shared helper: great-circle RMSE

```{r helpers}
gc_dist <- function(lon1, lat1, lon2, lat2) {
  r <- pi / 180
  acos(pmin(1, sin(lat1 * r) * sin(lat2 * r) +
              cos(lat1 * r) * cos(lat2 * r) * cos((lon2 - lon1) * r))) * 6371
}
rmse_km <- function(fit_obj) {
  sqrt(mean(gc_dist(fit_obj$fit$lon, fit_obj$fit$lat, true_lon, true_lat)^2))
}
```

---

## Model 1: Light only

```{r fit_lo, message = FALSE}
fit_lo <- TwilightFreeGrid(
  date_time         = times,
  light             = obs,
  grid              = grid,
  calibration       = cal,
  likelihood_params = lp,
  step_hours        = 12,
  diffusion         = 200
)
cat(sprintf("Light-only  ŌĆö mean lat: %+.1f┬░   RMSE: %d km\n",
            mean(fit_lo$fit$lat), round(rmse_km(fit_lo))))
```

The MAP is at approximately +24┬░: wrong hemisphere, ~5 400 km from the true track.

---

## Model 2: Hemisphere prior

`hemisphere_prior("S")` adds log(10Ōü╗┬│) Ōēł ŌłÆ6.9 nats to every northern-hemisphere cell.
That is strong enough to break the equinox tie but weak enough to be overridden by
decisive light data outside the equinox window.

```{r fit_hemi, message = FALSE}
term_hemi <- location_term(
  "hemisphere",
  source = hemisphere_prior(function(d) "S", softness = 1e-3),
  rule   = identity_rule()
)

fit_hemi <- TwilightFreeGrid(
  date_time         = times,
  light             = obs,
  grid              = grid,
  calibration       = cal,
  likelihood_params = lp,
  step_hours        = 12,
  diffusion         = 200,
  terms             = list(term_hemi)
)
cat(sprintf("+ hemi prior ŌĆö mean lat: %+.1f┬░   RMSE: %d km\n",
            mean(fit_hemi$fit$lat), round(rmse_km(fit_hemi))))
```

---

## Model 3: Bathymetry floor

The tag records a maximum dive depth of **500 m** at every knot.  Water depths of
200 m (northern subtropical shelf-like region in this synthetic scenario) are too
shallow for a 500 m dive and receive ŌłÆŌł× log-likelihood from `floor_rule()`.

```{r fit_bathy, message = FALSE}
# Synthetic depth raster: northern cells (lat > 0) = 200 m; southern (lat < 0) = 4 000 m.
# raster() fills rows top-to-bottom, so the first matrix row maps to the northern band.
depth_r <- raster::raster(
  matrix(c(200, 4000), ncol = 1),
  xmn = 145, xmx = 155, ymn = -35, ymx = 35
)

dive_df    <- data.frame(time = times, value = 500)
term_bathy <- location_term(
  "bathy",
  tag    = dive_df,
  reduce = "max",
  source = bathy_source(rast = depth_r),
  rule   = floor_rule()
)

fit_bathy <- TwilightFreeGrid(
  date_time         = times,
  light             = obs,
  grid              = grid,
  calibration       = cal,
  likelihood_params = lp,
  step_hours        = 12,
  diffusion         = 200,
  terms             = list(term_bathy)
)
cat(sprintf("+ bathy      ŌĆö mean lat: %+.1f┬░   RMSE: %d km\n",
            mean(fit_bathy$fit$lat), round(rmse_km(fit_bathy))))
```

Bathymetry alone corrects the hemisphere because the 500 m dive is physically impossible
in 200 m water.  The two constraints (prior and bathy) use entirely different data
streams ŌĆö breeding-range knowledge versus dive behaviour ŌĆö and each independently
resolves the ambiguity.

---

## Model 4: All three terms combined

We add a sea surface temperature channel via `function_source()`.  The tag records
~14 ┬░C; the synthetic SST field is warmest in the northern subtropics, matching the
true (southern) location best.

```{r fit_full, message = FALSE}
sst_fn <- function_source(function(lon, lat, date) {
  14 + (lat - true_lat) * 0.3     # 14 ┬░C at true_lat; warmer to the north
})
sst_tag  <- data.frame(time = times, value = 14)
term_sst <- location_term(
  "sst",
  tag    = sst_tag,
  reduce = "mean",
  source = sst_fn,
  rule   = student_rule(sd = 3, df = 4)
)

fit_full <- TwilightFreeGrid(
  date_time         = times,
  light             = obs,
  grid              = grid,
  calibration       = cal,
  likelihood_params = lp,
  step_hours        = 12,
  diffusion         = 200,
  terms             = list(term_hemi, term_bathy, term_sst)
)
cat(sprintf("+ all three  ŌĆö mean lat: %+.1f┬░   RMSE: %d km\n",
            mean(fit_full$fit$lat), round(rmse_km(fit_full))))
```

---

## Comparing the models

```{r compare_plot, echo = FALSE, fig.cap = "Posterior latitude at each 12-hour knot (5 knots over 2 days). Dashed line = true latitude (ŌłÆ25┬░). Every constraint-augmented model recovers the southern hemisphere; the light-only MAP sits at the northern mirror (~+24┬░)."}
k_steps  <- length(fit_lo$fit$lat)
knot_t   <- seq(min(times), max(times), length.out = k_steps)

df_all <- rbind(
  data.frame(time  = knot_t, lat = fit_lo$fit$lat,    model = "Light only"),
  data.frame(time  = knot_t, lat = fit_hemi$fit$lat,  model = "Light + prior"),
  data.frame(time  = knot_t, lat = fit_bathy$fit$lat, model = "Light + bathy"),
  data.frame(time  = knot_t, lat = fit_full$fit$lat,  model = "Light + all three")
)
df_all$model <- factor(df_all$model,
                       levels = c("Light only", "Light + prior",
                                  "Light + bathy", "Light + all three"))

ggplot2::ggplot(df_all, ggplot2::aes(x = time, y = lat, colour = model)) +
  ggplot2::geom_line(linewidth = 1.0) +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::geom_hline(yintercept = true_lat, linetype = "dashed",
                      colour = "black", linewidth = 0.6) +
  ggplot2::annotate("text",
                    x     = min(knot_t),
                    y     = true_lat - 1.5,
                    label = paste0("true lat = ", true_lat, "┬░"),
                    hjust = 0, size = 3.2, colour = "black") +
  ggplot2::scale_colour_manual(
    values = c("Light only"        = "#e74c3c",
               "Light + prior"     = "#2980b9",
               "Light + bathy"     = "#27ae60",
               "Light + all three" = "#8e44ad")
  ) +
  ggplot2::scale_x_datetime(date_labels = "%b %d") +
  ggplot2::labs(x = NULL, y = "Posterior latitude (┬░)", colour = NULL) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(legend.position = "bottom")
```

### RMSE table

```{r rmse_table, echo = FALSE}
knitr::kable(
  data.frame(
    Model   = c("Light only", "Light + prior", "Light + bathy", "Light + all three"),
    RMSE_km = round(c(rmse_km(fit_lo), rmse_km(fit_hemi),
                      rmse_km(fit_bathy), rmse_km(fit_full)))
  ),
  caption   = "RMSE vs. the known true track (km).",
  col.names = c("Model", "RMSE (km)")
)
```

Both the hemisphere prior and the bathymetry floor reduce RMSE by ~29├Ś independently.
Their combination, plus the SST term, provides redundancy: if one data stream is
absent or unreliable, the others continue to anchor the track.

---

## Building a bespoke sensor with `function_source()` and `custom_rule()`

Any callable `fn(lon, lat, date)` can be promoted to a `tf_source`, and any
`fn(obs, expected)` can be promoted to a `tf_rule`.  Here we sketch a
**Poisson acoustic detection count** rule, where the tag logs how many detections
it picks up from a fixed hydrophone mooring and the expected count decays
exponentially with distance.

```{r custom_demo, eval = FALSE}
# ---------- source ----------------------------------------------------
mooring_lon <- 152; mooring_lat <- -28

detection_source <- function_source(function(lon, lat, date) {
  dist_km <- gc_dist(lon, lat, mooring_lon, mooring_lat)
  100 * exp(-dist_km / 300)   # expected detections; 300 km scale length
})

# ---------- rule ------------------------------------------------------
poisson_rule <- custom_rule(function(obs, expected) {
  lam <- pmax(expected, 1e-6)        # guard log(0)
  obs * log(lam) - lam               # Poisson log-likelihood (omits log(obs!))
})

# ---------- term ------------------------------------------------------
acoustic_tag <- data.frame(
  time  = times,
  value = rpois(length(times), lambda = 40)  # ~40 detections / 10 min
)
term_acoustic <- location_term(
  "acoustic",
  tag    = acoustic_tag,
  reduce = "sum",            # total detections per 12 h knot
  source = detection_source,
  rule   = poisson_rule,
  weight = 0.5               # down-weight relative to light
)

# Add to any TwilightFreeGrid() call:
fit_acoustic <- TwilightFreeGrid(
  date_time = times, light = obs, grid = grid,
  calibration = cal, likelihood_params = lp,
  step_hours = 12, diffusion = 200,
  terms = list(term_hemi, term_acoustic)
)
```

The same pattern works for geomagnetic inclination (IGRF model wrapped in
`function_source()`), tidal-phase constraints, accelerometer activity budgets, or any
other channel that provides a scalar observation per knot and a predicted field over
the grid.

---

## Quick-reference: building blocks

| Function | Purpose |
|----------|---------|
| `location_term()` | Bundle tag channel + source + rule into one additive term |
| `hemisphere_prior()` | Soft/hard hemispheric prior, keyed by calendar date |
| `bathy_source()` | Seafloor depth (ETOPO or user raster) for dive constraints |
| `sst_source()` | Daily NOAA OISST (or user raster) for temperature matching |
| `raster_source()` | Any static or date-indexed `RasterLayer` as a source |
| `function_source()` | Promote `fn(lon, lat, date)` to a `tf_source` |
| `floor_rule()` | One-sided hard/soft constraint (dive depth, altitude, terrain) |
| `student_rule()` | Robust symmetric match (SST, salinity) |
| `gaussian_rule()` | Symmetric squared-error match |
| `ceiling_rule()` | Mirror of `floor_rule()`: upper-bound constraint |
| `mask_rule()` | Binary permit/forbid mask (sea vs. land) |
| `custom_rule()` | Bring-your-own `function(obs, expected)` rule |
| `identity_rule()` | Pass-through for sources already in log-density form |
