---
title: "Northern elephant seals: twilight-free geolocation with and without sensor fusion"
subtitle: "Ten post-moult TDR Mk9 deployments validated against Argos"
author: "Aidan D. Bindoff"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M %Z')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    df_print: paged
    fig_width: 9
    fig_height: 6
params:
  # ---- where the data live -------------------------------------------------
  # A local copy of the co-author's Google Drive folder. These data are not
  # publicly shareable: the directory is gitignored and .Rbuildignored.
  data_dir:   "data/nes_untracked"
  cache_dir:  "analysis/cache/nes"
  out_dir:    "analysis/output/nes"

  # ---- ingest --------------------------------------------------------------
  decimate_minutes:   30      # the archives are 4-second; thin to this window (see 3.3)
  surface_depth_m:     5      # at or above this depth a temperature reading is "surface"
  clock_search_min:  120      # half-width of the clock offset search, minutes
  clock_max_accept_min: 20    # reject an offset OR a drift larger than this (see 3.7)
  estimate_clock:   TRUE      # estimate the per-tag clock offset from the known endpoints
  apply_clock_correction: FALSE   # ...but do NOT shift by it; see 3.7 for why
  clock_hires_minutes: 1      # resolution of the light used for the clock (see 3.7)
  clock_window_days:   3      # days of light around each known fix
  cal_window_days:    15      # days at the release site used to fit the light response

  # ---- ground truth --------------------------------------------------------
  argos_keep_lc: "3,2,1,0,A,B"
  argos_vmax_kmh:     10      # sustained NES travel is 3-4 km/h
  argos_max_gap_h:    24      # do not interpolate truth across longer gaps

  # ---- models --------------------------------------------------------------
  step_hours:         12
  seed:               42
  # hierarchical
  sweeps:           4000
  burn:             1500
  thin:                5
  mesh_pad_lon:       74      # deploy is a single point, so padding sets the mesh
  mesh_pad_lat:       26
  coarse_res:          2
  surrogate_diffusion: 110    # km/sqrt(day), proposal only
  # grid HMM
  grid_cell_deg:       1
  grid_diffusion:    110
  # FFBS
  ffbs_n:              3
  ffbs_particles:   1000

  # ---- sensor-fusion terms -------------------------------------------------
  use_sst:         FALSE      # Mk9 temperature as an SST proxy; see 3.6 for why not
  sst_depth_band: "2,10"      # m; the band in which the thermistor tracks SST
  sst_sd:            1.5      # student_rule scale, deg C
  sst_df:              4
  bathy_margin_m:    300      # subtract from the knot's deepest dive before constraining
  bathy_sd:          300      # soft floor_rule tolerance
  soft_mask_penalty: -20

  # ---- environmental layers ------------------------------------------------
  fetch_env:        TRUE
  bathy_stride:        4      # ETOPO1 is 1 arc-min; 4 gives ~7 km

  # ---- limits (set for a fast smoke test) ----------------------------------
  # `null`, not NA: YAML turns a bare NA into the *string* "NA", which passes
  # is.na() and silently corrupts head()/min() downstream.
  max_tags:         null
  max_days:         null
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE,
                      collapse = FALSE, comment = "#>", dpi = 110)
options(stringsAsFactors = FALSE, width = 115)
Sys.setlocale("LC_TIME", "C")   # the archives use "21:02:00 22-May-2021"
set.seed(params$seed)

# The path parameters are written relative to the PACKAGE ROOT, but knitr's
# working directory defaults to the directory holding this file, so rendering it
# any ordinary way ("Knit" in RStudio, or rmarkdown::render() with no
# `knit_root_dir`) looks for the data one level too deep and stops at the
# environment check. Find the root by walking up for DESCRIPTION and point knitr
# at it. `params` itself cannot be rewritten here, since knitr locks it, and
# root.dir takes effect from the next chunk on, which is why this is done in
# `setup` and nowhere later.
knitr::opts_knit$set(root.dir = local({
  d <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
  repeat {
    if (file.exists(file.path(d, "DESCRIPTION"))) return(d)
    up <- dirname(d)
    if (identical(up, d)) return(normalizePath(getwd(), winslash = "/"))
    d <- up
  }
}))
```

# What this document does

Ten female northern elephant seals (*Mirounga angustirostris*) were tagged at
Ano Nuevo in May-June 2021 with a Wildlife Computers **Mk9** archival TDR
(4-second light, depth and external temperature) and a **SPOT** Argos
transmitter, and recovered after post-moult foraging trips of 137 to 248 days.
The Argos track is independent of the light record, so it is ground truth for a
light-only reconstruction.

The document reconstructs each track from **raw light alone** and from **light
plus the tag's own auxiliary channels**, in three engines, and scores every
reconstruction against Argos. It answers four questions.

1. How accurate is `invTwilightFree` on real, dive-wrecked light data?
2. What does sensor fusion add? The tag's depth channel constrains position
   through bathymetry; its temperature channel constrains position through sea
   surface temperature. Both are free, in the sense that they are already on the
   tag.
3. **Does sensor fusion make the fit faster?** It should, and section 8 measures
   it. A bathymetric map rules out a candidate location *outright* once the
   animal has dived past that seabed, and an outright exclusion is cheaper than
   an improbability: the engine can skip the light likelihood there entirely.
   That is only true of a **hard** constraint; a soft one still has to be
   evaluated everywhere. Both are run, and the difference is the point.
4. Does correcting the tag clock help? Longitude error is clock error, and these
   deployments run for eight months.

## The batches

| Key | Engine | Terms | Tags | Clock |
|---|---|---|---|---|
| `hier_light` | `TwilightFreeHier` | none | all | as estimated (none accepted) |
| `hier_fusion` | `TwilightFreeHier` | soft bathy + SST + soft mask | all | as estimated (none accepted) |
| `hier_light_forced` | `TwilightFreeHier` | none | all | **forced, rule overridden** |
| `grid_light` | `TwilightFreeGrid` | none | all | as estimated (none accepted) |
| `grid_fusion` | `TwilightFreeGrid` | **hard** bathy + hard mask + SST | all | as estimated (none accepted) |
| `ffbs_light` | `TwilightFreeSMC("ffbs")` | none | subset | as estimated (none accepted) |
| `ffbs_fusion` | `TwilightFreeSMC("ffbs")` | soft bathy + SST | subset | as estimated (none accepted) |

`hier_light` and `hier_light_forced` differ only in whether the clock estimates
of section 3.7 were applied in defiance of the acceptance rule, which turns "is
this estimator safe to trust?" into a measurement rather than an opinion. No
estimate passes the rule, so every other batch runs on raw timestamps.

## Longitude convention

Two seals (PTT 214397 and 214402) cross the antimeridian. **Everything here uses
0-360 degrees east**, so those tracks stay contiguous and no great-circle step is
corrupted by a 360-degree wrap. Ano Nuevo is therefore 237.67, not -122.33. Both
environmental products are fetched from their native 0-360 ERDDAP datasets
(`ncdcOisst21Agg`, `etopo360`), so nothing is reprojected anywhere.

# Environment and provenance

```{r env-check}
need <- c("invTwilightFree", "terra", "data.table", "ncdf4", "xml2")
missing <- need[!vapply(need, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) stop("missing packages: ", paste(missing, collapse = ", "))
suppressPackageStartupMessages({ library(invTwilightFree); library(terra); library(data.table) })

cat("R               :", R.version.string, "\n")
for (p in need) cat(sprintf("%-16s: %s\n", p, as.character(packageVersion(p))))
dir.create(params$cache_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(params$out_dir,   recursive = TRUE, showWarnings = FALSE)
if (!dir.exists(params$data_dir)) stop("params$data_dir not found: ", params$data_dir)
```

```{r manifest}
# Every file that feeds the analysis, hashed, so the results are tied to bytes
# rather than to a folder name that might be re-synced later.
data_files <- list.files(params$data_dir, recursive = TRUE, full.names = TRUE)
manifest <- data.frame(
  path  = sub(paste0("^", normalizePath(params$data_dir, winslash = "/"), "/?"), "",
              normalizePath(data_files, winslash = "/")),
  bytes = file.size(data_files),
  md5   = unname(tools::md5sum(data_files)), row.names = NULL)
write.csv(manifest, file.path(params$out_dir, "data_manifest.csv"), row.names = FALSE)
cat(nrow(manifest), "files,", sprintf("%.2f GB\n", sum(manifest$bytes) / 1024^3))
head(manifest[order(-manifest$bytes), ], 8)
```

# Helpers

Everything beyond `invTwilightFree`, `terra`, `data.table`, `ncdf4` and `xml2`
is defined here, so there are no hidden dependencies.

```{r helpers}
# ---- geometry ----------------------------------------------------------------
lon360 <- function(x) (x %% 360 + 360) %% 360

gc_dist_km <- function(lon1, lat1, lon2, lat2, R = 6371.0088) {
  d2r <- pi / 180
  a <- sin((lat2 - lat1) * d2r / 2)^2 +
       cos(lat1 * d2r) * cos(lat2 * d2r) * sin((lon2 - lon1) * d2r / 2)^2
  2 * R * asin(pmin(1, sqrt(a)))
}

# Signed longitude difference, shortest way round. Needed for bias and RMSE in
# longitude: a raw subtraction across the dateline reads as 360 degrees of error.
dlon <- function(a, b) ((a - b + 180) %% 360) - 180

# ---- io ----------------------------------------------------------------------
# The archives use "%H:%M:%S %d-%b-%Y"; LC_TIME is forced to C in setup so %b
# parses regardless of the machine's locale.
parse_wc_time <- function(x) as.POSIXct(x, format = "%H:%M:%S %d-%b-%Y", tz = "UTC")

# Minimal xlsx reader: an xlsx is a zip of XML, and readxl segfaults in this R
# build. Returns a character matrix of the used range of one worksheet.
read_xlsx_grid <- function(path, sheet = 1L) {
  ns <- c(d = "http://schemas.openxmlformats.org/spreadsheetml/2006/main")
  tmp <- file.path(tempdir(), "xlsx"); unlink(tmp, recursive = TRUE)
  utils::unzip(path, exdir = tmp)
  ss <- character(0)
  sfile <- file.path(tmp, "xl", "sharedStrings.xml")
  if (file.exists(sfile)) {
    ss <- vapply(xml2::xml_find_all(xml2::read_xml(sfile), "d:si", ns),
                 function(si) paste0(xml2::xml_text(xml2::xml_find_all(si, ".//d:t", ns)),
                                     collapse = ""), "")
  }
  ws <- xml2::read_xml(file.path(tmp, "xl", "worksheets", sprintf("sheet%d.xml", sheet)))
  rows <- xml2::xml_find_all(ws, "//d:row", ns)
  cell_col <- function(ref) {                       # "BC12" -> 55
    L <- utf8ToInt(toupper(gsub("[0-9]", "", ref))) - 64L
    sum(L * 26^rev(seq_along(L) - 1))
  }
  parsed <- Filter(Negate(is.null), lapply(rows, function(r) {
    cs <- xml2::xml_find_all(r, "d:c", ns)
    if (!length(cs)) return(NULL)
    list(idx = vapply(xml2::xml_attr(cs, "r"), cell_col, 1),
         val = vapply(seq_along(cs), function(i) {
           cc <- cs[[i]]; ty <- xml2::xml_attr(cc, "t")
           v <- xml2::xml_find_first(cc, "d:v", ns)
           if (is.na(v)) {
             is_el <- xml2::xml_find_first(cc, "d:is", ns)
             if (is.na(is_el)) "" else xml2::xml_text(is_el)
           } else if (identical(ty, "s")) ss[as.integer(xml2::xml_text(v)) + 1L]
           else xml2::xml_text(v)
         }, ""))
  }))
  ncol <- max(unlist(lapply(parsed, function(p) p$idx)))
  out <- matrix("", nrow = length(parsed), ncol = ncol)
  for (i in seq_along(parsed)) out[i, parsed[[i]]$idx] <- parsed[[i]]$val
  out
}

# ---- misc --------------------------------------------------------------------
# Thin a 4-second dive record to one row per `minutes`, taking from each window
# the quantities the model actually needs (see 3.3 for why this is not a
# subsample):
#   light      max, i.e. the least-attenuated sample in the window
#   depth_max  deepest point, which is what the bathymetry floor constrains
#   depth_min  shallowest point, which says whether the animal surfaced at all
#   temp_surf  temperature at the shallowest record, the most SST-like reading
# The timestamp is the instant the retained light value was actually measured,
# NOT the start of the window. Labelling a window with its first timestamp while
# keeping the maximum light from anywhere inside it stamps every value about half
# a window early, which is a pure longitude bias of (minutes/2)/4 degrees --
# identical for every tag, and easily mistaken for a fleet of clock errors.
# Measured on these data: +15.0 min at a 30-minute window and +30.0 min at 60,
# falling to about 1 minute once the timestamp follows the sample.
decimate_dive <- function(d, minutes) {
  d[, .b := floor(as.numeric(time) / (minutes * 60))]
  out <- d[, {
    j <- which.max(light)                    # the sample the light value comes from
    i <- which.min(depth)                    # the shallowest point, for temperature
    .(time = time[j],
      light = as.numeric(light[j]),
      depth_max = as.numeric(max(depth, na.rm = TRUE)),
      depth_min = as.numeric(min(depth, na.rm = TRUE)),
      temp_surf = if (length(i)) as.numeric(temp[i]) else NA_real_)
  }, by = .b]
  out[, .b := NULL]
  setorder(out, time)
  as.data.frame(out[is.finite(light)])
}

# Linear interpolation of an Argos track to arbitrary times, along the shortest
# arc in longitude. Used both for the SST validation and for scoring.
interp_argos_xy <- function(a, times, max_gap_h = 24) {
  at <- as.numeric(a$time); qt <- as.numeric(times)
  j <- findInterval(qt, at); ok <- j >= 1 & j < length(at)
  lo <- la <- rep(NA_real_, length(qt)); jj <- j[ok]
  good <- (at[jj + 1] - at[jj]) / 3600 <= max_gap_h
  w <- (qt[ok] - at[jj]) / (at[jj + 1] - at[jj]); idx <- which(ok)[good]
  lo[idx] <- lon360(a$lon[jj[good]] + w[good] * dlon(a$lon[jj[good] + 1], a$lon[jj[good]]))
  la[idx] <- a$lat[jj[good]] + w[good] * (a$lat[jj[good] + 1] - a$lat[jj[good]])
  list(lon = lo, lat = la)
}

timed <- function(expr) {
  t0 <- proc.time()[["elapsed"]]
  val <- force(expr)
  list(value = val, seconds = proc.time()[["elapsed"]] - t0)
}

# Cache a plain-R result (never a terra object: those are external pointers).
# NB: never write return() inside the expression. It is forced as a promise
# inside cache_rds(), so a return() there unwinds the wrong frame and the
# assignment silently never happens.
#
# The key is the only thing tying a cached result to the code that made it, so
# BUMP THE SUFFIX whenever anything upstream of a fit changes. The `_v2` keys
# below date from the change to how the light response is calibrated (5.2),
# which moves every fit and every clock offset; a run against `_v1` caches would
# have reported the old numbers under the new text without a word of complaint.
cache_rds <- function(key, expr, dir = params$cache_dir, refresh = FALSE) {
  f <- file.path(dir, paste0(key, ".rds"))
  if (!refresh && file.exists(f)) { message("cache hit: ", key); return(readRDS(f)) }
  val <- force(expr); saveRDS(val, f); val
}
```

# Ingest and quality control

## The deployment table

The spreadsheet is the authoritative crosswalk between the three identifiers
that appear in the data: the TOPP id (which names the archive file), the Mk9
serial (which also names the archive file) and the Argos PTT (which names the
Argos folder). Dates are Excel serials.

```{r meta}
xlsx <- list.files(params$data_dir, pattern = "\\.xlsx$", full.names = TRUE)[1]
g <- read_xlsx_grid(xlsx)
tbl <- as.data.frame(g[-(1:2), , drop = FALSE], stringsAsFactors = FALSE)
names(tbl) <- make.names(g[2, ], unique = TRUE)     # row 1 is a merged banner

meta <- data.frame(
  id      = tbl$TOPP.ID,
  season  = tbl$Season,
  seal    = tbl$Seal.ID,
  age     = suppressWarnings(as.numeric(tbl$Deployment.Age)),
  sex     = tbl$Sex,
  ptt     = tbl$Sat.Tag.PTT,
  tdr     = tbl$TDR..,
  deploy_date  = as.Date(suppressWarnings(as.numeric(tbl$Date)),   origin = "1899-12-30"),
  recover_date = as.Date(suppressWarnings(as.numeric(tbl$Date.1)), origin = "1899-12-30"),
  stringsAsFactors = FALSE)
meta <- meta[!is.na(meta$id) & nzchar(meta$id), ]
meta$days <- as.numeric(meta$recover_date - meta$deploy_date)
meta
```

## Archival records

The Mk9 files are named `<TOPP id>_<Mk9 serial>.csv` and carry `Time`, `Depth`,
`External Temp` and `Light Level` at a 4-second cadence: 3 to 5 million rows and
130 to 240 MB per animal. One file in the folder (`2021030_19A1005.csv`, 1.5 GB)
is a different instrument altogether, with `Wet/Dry`, accelerometry,
conductivity and salinity but **no light channel**, so it cannot be geolocated
by this method. It is dropped, with the reason recorded rather than silently
skipped.

```{r archive-list}
LIGHT_COLS <- c("Time", "Depth", "External Temp", "Light Level")

archive_paths <- list.files(params$data_dir, pattern = "^[0-9]+_.*\\.csv$", full.names = TRUE)
archive_hdr <- lapply(archive_paths, function(f) names(fread(f, nrows = 0L)))
has_light <- vapply(archive_hdr, function(h) all(LIGHT_COLS %in% h), logical(1))

cat("archive files:", length(archive_paths), "\n")
for (i in seq_along(archive_paths))
  cat(sprintf("  %-26s %6.0f MB  %s\n", basename(archive_paths[i]),
              file.size(archive_paths[i]) / 1024^2,
              if (has_light[i]) "light+depth+temp" else
                paste("DROPPED, no light channel; columns:",
                      paste(utils::head(archive_hdr[[i]], 6), collapse = ","))))
archive_paths <- archive_paths[has_light]
```

Thinning to `decimate_minutes` happens at read time, because the likelihood cost
is linear in the number of observations. **How** it is thinned turns out to
matter more than by how much, and getting it wrong is the single largest error
available in this analysis, so it is justified rather than assumed.

A northern elephant seal dives more or less continuously. An instantaneous light
sample therefore records the animal's dive state, not the light environment: it
is dark most of the time regardless of the sun. Subsampling the record, which is
the obvious thing to do to a 4-second series, samples that dive state. Taking
the **maximum** over a window instead returns the least-attenuated sample in it,
which is the one taken nearest the surface, and that is the light envelope the
solar-geometry model is about.

The window has to be at least as long as a dive cycle, or some windows contain
no surfacing at all. The chunk below measures the dive cycle from the data and
scores each candidate rule by its correlation with the solar zenith angle
computed at the animal's **known** position, which is the sharpest available
test of whether a positional signal survives.

```{r decimation-choice}
COLONY_LON <- 237.67; COLONY_LAT <- 37.11   # Ano Nuevo, 0-360 convention
probe <- fread(archive_paths[1], showProgress = FALSE,
               select = c("Time", "Depth", "Light Level"))
setnames(probe, c("time", "depth", "light"))
probe[, time := parse_wc_time(time)]
probe <- probe[!is.na(time) & !is.na(light)]
# Mid-trip, where the animal is diving hardest. The reference zenith is computed
# at the colony, so this is a lower bound on the achievable correlation: the
# animal has moved, which can only weaken it.
mid <- probe[time >= min(time) + 60 * 86400 & time < min(time) + 70 * 86400]

gaps <- as.numeric(diff(mid[depth <= params$surface_depth_m, time]))
gaps <- gaps[gaps > 60]
cat(sprintf("surface time: %.1f%% of records | gap between surfacings (min): median %.1f, q99 %.1f\n",
            100 * mean(mid$depth <= params$surface_depth_m), median(gaps) / 60,
            quantile(gaps, .99) / 60))

score_rule <- function(mins, fun, label) {
  agg <- mid[, .(v = as.numeric(fun(light)), t = as.numeric(time[1])),
             by = .(b = floor(as.numeric(time) / (mins * 60)))][is.finite(v)]
  data.frame(rule = label, minutes = mins, n = nrow(agg),
             cor_with_zenith = round(cor(agg$v, -solar_zenith(
               agg$t, rep(COLONY_LON, nrow(agg)),
               rep(COLONY_LAT, nrow(agg)))), 3))
}
rules <- do.call(rbind, c(
  lapply(c(10, 30, 60), function(m) score_rule(m, function(v) v[1], "subsample (first)")),
  lapply(c(5, 10, 20, 30, 60), function(m) score_rule(m, max, "maximum"))))
print(rules, row.names = FALSE)
```

The maximum rule roughly doubles the correlation, and it plateaus once the
window exceeds the dive cycle; a longer window buys nothing and costs temporal
resolution. `decimate_minutes = 30` sits at that plateau. At a 12-hour knot
spacing it still puts 24 observations in every knot, and it is objective and
automatic, so it does not reintroduce the manual editing step the method exists
to remove.

```{r archives}
# The cache key carries a format version: this chunk returns a list of
# (main, ends) pairs, and a cache written by an earlier layout would be read
# back as NULL for both.
archives <- cache_rds(sprintf("archives_v3_%dmin", params$decimate_minutes), {
  p <- archive_paths
  if (!is.null(params$max_tags)) p <- utils::head(p, params$max_tags)
  out <- lapply(p, function(f) {
    message("reading ", basename(f))
    d <- fread(f, select = LIGHT_COLS, showProgress = FALSE)
    setnames(d, c("time", "depth", "temp", "light"))
    d[, time := parse_wc_time(time)]
    d <- d[!is.na(time) & !is.na(light)]
    setorder(d, time)
    d <- d[!duplicated(time)]
    # Two products from the one expensive read: the working series, and a
    # high-resolution slice around each endpoint for the clock estimate, which
    # cannot resolve a twilight more finely than the sampling interval (3.7).
    w <- params$clock_window_days * 86400
    ends <- d[time <= min(time) + w | time >= max(time) - w]
    list(main = decimate_dive(d, params$decimate_minutes),
         ends = decimate_dive(ends, params$clock_hires_minutes))
  })
  names(out) <- sub("_.*$", "", basename(p))         # TOPP id
  out
})
archive_ends <- lapply(archives, `[[`, "ends")
archives     <- lapply(archives, `[[`, "main")

archive_summary <- do.call(rbind, lapply(names(archives), function(id) {
  d <- archives[[id]]
  data.frame(id = id, n = nrow(d), start = min(d$time), end = max(d$time),
             days = round(as.numeric(difftime(max(d$time), min(d$time), units = "days")), 1),
             light_q05 = quantile(d$light, .05), light_q95 = quantile(d$light, .95),
             light_max = max(d$light),
             depth_med = round(median(d$depth_max, na.rm = TRUE)),
             depth_max = round(max(d$depth_max, na.rm = TRUE)),
             pct_windows_surfaced = round(100 * mean(d$depth_min <= params$surface_depth_m,
                                                     na.rm = TRUE), 1),
             row.names = NULL)
}))
archive_summary
```

`depth_med` is the headline number for how hostile these data are: the typical
30-minute window reaches several hundred metres, where the light sensor sees
almost nothing. `pct_windows_surfaced` is the reassurance that the decimation
rule works: it is the percentage of windows in which the animal came within
`surface_depth_m` of the surface at least once, and therefore the percentage for
which the window maximum is a near-surface reading.

## Argos

Each PTT folder holds a Wildlife Computers data package. Two files matter:
`-Locations.csv` (the track) and `-RTC.csv` (the SPOT tag's own clock-drift
record, used as context in section 3.7).

The `Locations` file has two kinds of row. `Type == "Argos"` rows are satellite
fixes with a location class, and are the ground truth. `Type == "User"` rows are
a single researcher-entered position per animal, 37.11 N 122.33 W, dated to the
deployment day. **That is the deployment location** the metadata otherwise
lacks. It is used to pin the first knot, and excluded from the ground-truth
comparison, because it is an input rather than an independent observation.

```{r argos-unpack}
# The Argos packages arrive as a zip of per-PTT zips.
zips <- list.files(params$data_dir, pattern = "^drive-download.*\\.zip$", full.names = TRUE)
argos_dir <- file.path(params$data_dir, "argos")
if (length(zips) && !dir.exists(argos_dir)) {
  dir.create(argos_dir, showWarnings = FALSE)
  utils::unzip(zips[1], exdir = argos_dir)
  for (z in list.files(argos_dir, pattern = "\\.zip$", full.names = TRUE))
    # One package (214393) has a corrupt MinMaxDepth entry we do not need, so
    # this warns and continues rather than stopping.
    suppressWarnings(try(utils::unzip(z, exdir = file.path(argos_dir, sub("\\.zip$", "", basename(z)))),
                         silent = TRUE))
}
loc_files <- list.files(argos_dir, pattern = "-Locations\\.csv$", recursive = TRUE, full.names = TRUE)
cat("Argos location files:", length(loc_files), "\n")
```

```{r argos-read}
A <- rbindlist(lapply(loc_files, function(f)
  fread(f, showProgress = FALSE, colClasses = list(character = c("Ptt", "Quality")))), fill = TRUE)
A[, time := parse_wc_time(Date)]
A[, lon := lon360(Longitude)][, lat := Latitude]

deploy_fix <- A[Type == "User", .(deploy_lon = median(lon), deploy_lat = median(lat),
                                  user_date = min(as.Date(time))), by = Ptt]
cat("deployment fixes (Type == 'User'):\n"); print(deploy_fix)

keep_lc <- trimws(strsplit(params$argos_keep_lc, ",")[[1]])
A <- A[Type == "Argos" & Quality %in% keep_lc & !is.na(time) & !is.na(lon)]

# Forward speed filter. Elephant seals sustain 3-4 km/h; a step implying more
# than argos_vmax_kmh from the last accepted fix is a bad Argos solution rather
# than a fast seal.
speed_filter <- function(d, vmax) {
  setorder(d, time)
  if (nrow(d) < 3) return(d)
  keep <- rep(TRUE, nrow(d)); last <- 1L
  for (i in 2:nrow(d)) {
    dt <- as.numeric(difftime(d$time[i], d$time[last], units = "hours"))
    if (dt <= 0) { keep[i] <- FALSE; next }
    if (gc_dist_km(d$lon[last], d$lat[last], d$lon[i], d$lat[i]) / dt > vmax) keep[i] <- FALSE
    else last <- i
  }
  d[keep]
}
n_before <- nrow(A)
argos <- lapply(split(A, A$Ptt), speed_filter, vmax = params$argos_vmax_kmh)
cat(sprintf("speed filter removed %d of %d fixes (%.1f%%)\n",
            n_before - sum(vapply(argos, nrow, 0L)), n_before,
            100 * (1 - sum(vapply(argos, nrow, 0L)) / n_before)))

argos_summary <- do.call(rbind, lapply(names(argos), function(p) {
  d <- argos[[p]]
  data.frame(ptt = p, n = nrow(d),
             per_day = round(nrow(d) / as.numeric(difftime(max(d$time), min(d$time),
                                                           units = "days")), 1),
             lon_min = round(min(d$lon), 1), lon_max = round(max(d$lon), 1),
             lat_min = round(min(d$lat), 1), lat_max = round(max(d$lat), 1), row.names = NULL)
}))
argos_summary
```

## Assembling the panel

```{r tags}

build_tag <- function(id) {
  m <- meta[match(id, meta$id), ]
  if (is.na(m$ptt) || !m$ptt %in% names(argos)) return(NULL)
  a <- argos[[m$ptt]]
  d <- archives[[id]]
  df <- deploy_fix[Ptt == m$ptt]
  if (!nrow(df)) return(NULL)

  # Window: the animal is at sea between the deployment day and the last Argos
  # fix. The archive starts when the Mk9 was switched on, before the animal
  # leaves the beach, so take the later of the two starts. The spreadsheet
  # deployment date and the User fix disagree by a day or two on some animals,
  # so take the later of those as well.
  t0 <- max(min(d$time), as.POSIXct(max(m$deploy_date, df$user_date), tz = "UTC"))
  t1 <- min(max(d$time), max(a$time))
  if (!is.null(params$max_days)) t1 <- min(t1, t0 + params$max_days * 86400)
  d <- d[d$time >= t0 & d$time <= t1, , drop = FALSE]
  a <- a[a$time >= t0 & a$time <= t1]
  if (nrow(d) < 1000 || nrow(a) < 50) return(NULL)

  # Recovery: the animal hauls out back at the colony, so take the median of the
  # good fixes in the last 72 h rather than trusting a single last fix.
  tail_fix <- a[a$time >= max(a$time) - 72 * 3600]
  list(id = id, ptt = m$ptt, seal = m$seal, light = d, argos = as.data.frame(a),
       deploy  = c(lon = df$deploy_lon, lat = df$deploy_lat),
       recover = c(lon = median(tail_fix$lon), lat = median(tail_fix$lat)),
       t0 = t0, t1 = t1, days = as.numeric(difftime(t1, t0, units = "days")))
}

tags <- lapply(names(archives), build_tag)
names(tags) <- names(archives)
tags <- tags[!vapply(tags, is.null, logical(1))]
stopifnot(length(tags) > 0)

tag_table <- do.call(rbind, lapply(tags, function(g) data.frame(
  id = g$id, seal = g$seal, ptt = g$ptt, days = round(g$days, 1),
  n_light = nrow(g$light), n_argos = nrow(g$argos),
  deploy_lon = round(g$deploy[["lon"]], 3), deploy_lat = round(g$deploy[["lat"]], 3),
  recover_lon = round(g$recover[["lon"]], 3), recover_lat = round(g$recover[["lat"]], 3),
  max_depth = round(max(g$light$depth_max, na.rm = TRUE)), row.names = NULL)))
tag_table
cat(sprintf("\npanel: %d tags, %.0f tag-days, %s light observations\n",
            length(tags), sum(tag_table$days), format(sum(tag_table$n_light), big.mark = ",")))
```

## Clock offsets

Longitude error is clock error: one minute is 0.25 degrees, roughly 20 km at
45 N. Two independent lines of evidence are available.

**The SPOT tags' own drift.** Each Argos package carries an `-RTC.csv` comparing
the tag's clock with a known real time. These are the *SPOT* clocks, not the Mk9
clocks, so they correct nothing here, but they show what a clock of this vintage
does over an eight-month deployment, which is the prior for what the Mk9 is
likely doing.

```{r rtc}
rtc_files <- list.files(argos_dir, pattern = "-RTC[.]csv$", recursive = TRUE, full.names = TRUE)
RTC <- rbindlist(lapply(rtc_files, fread, showProgress = FALSE), fill = TRUE)
RTC[, `:=`(tag_t  = parse_wc_time(paste(TagTime, TagDate)),
           real_t = parse_wc_time(paste(RealTime, RealDate)))]
RTC[, drift_s := as.numeric(tag_t - real_t, units = "secs")]
spot_drift <- RTC[CorrectionType == "Strong",
                  .(n_corrections = .N, last_drift_min = round(drift_s[.N] / 60, 2),
                    lon_equiv_deg = round(drift_s[.N] / 60 * 0.25, 2)), by = Ptt]
cat("SPOT clock drift at the last correction (positive = tag ahead of real time):\n")
print(spot_drift)
```

**A light-based estimate.** `calibrate_clock_from_endpoints()` searches for the
offset that maximises the light likelihood at the known deployment and recovery
positions, both of which are known here.

That estimate is run on **high-resolution light, not the working series.** A
twilight cannot be located more precisely than the sampling interval, so the
estimator's own uncertainty scales with the decimation. Measured on these ten
deployments, the width of the 2-log-unit interval around the peak was 10.6
minutes at 1-minute resolution, 23 at 5, 33 at 10 and 55 at 30. Estimating the
clock from the 30-minute working series would therefore return an offset with a
+/- 27 minute uncertainty, which is +/- 7 degrees of longitude: worse than the
error it is meant to remove. The `ends` slice built in section 3.3 exists for
exactly this, and costs nothing extra because it comes out of the same read.

```{r clock-estimate}
clocks <- if (!isTRUE(params$estimate_clock)) NULL else cache_rds("clocks_v2", {
  lapply(tags, function(g) {
    e <- archive_ends[[g$id]]
    e <- e[e$time >= g$t0 & e$time <= g$t1, ]
    tryCatch(calibrate_clock_from_endpoints(
      date_time = e$time, light = e$light,
      start_lon = g$deploy[["lon"]],  start_lat = g$deploy[["lat"]],
      end_lon   = g$recover[["lon"]], end_lat   = g$recover[["lat"]],
      calibration = NULL, likelihood_params = NULL,
      window_days = params$clock_window_days,
      search = params$clock_search_min * 60, by = 30),
      error = function(e) { message("clock estimation failed for ", g$id, ": ",
                                    conditionMessage(e)); NULL })
  })
})

# An offset that lands on the edge of the search window is not an estimate, it
# is a failure to find an interior optimum: the profile is flat or monotone and
# the optimiser has simply run to the boundary. Applying one injects error
# rather than removing it, so such tags keep their raw timestamps.
clock_table <- NULL
if (!is.null(clocks)) {
  clock_table <- do.call(rbind, lapply(names(clocks), function(id) {
    cl <- clocks[[id]]
    if (is.null(cl)) return(data.frame(id = id, ptt = tags[[id]]$ptt,
      offset_deploy_min = NA_real_, offset_recover_min = NA_real_,
      drift_min_per_day = NA_real_, lon_bias_deg = NA_real_, accepted = FALSE))
    od <- cl$offset_deploy / 60; orr <- cl$offset_retrieve / 60
    days <- as.numeric(difftime(tags[[id]]$t1, tags[[id]]$t0, units = "days"))
    data.frame(id = id, ptt = tags[[id]]$ptt,
               offset_deploy_min = round(od, 2), offset_recover_min = round(orr, 2),
               drift_min_per_day = round((orr - od) / days, 4),
               lon_bias_deg = round(od * 0.25, 2),
               drift_total_min = round(orr - od, 2),
               # Two ways to fail. A large OFFSET means the profile ran to the
               # search bound rather than finding an interior optimum. A large
               # DRIFT means the two endpoint estimates disagree by more than a
               # clock of this vintage can plausibly wander, and it is the more
               # dangerous of the two because `correct()` interpolates linearly
               # between them and so spreads the disagreement over the whole
               # record. The SPOT table above sets the scale: real drift over a
               # whole deployment runs -1.4 to +13.3 minutes.
               accepted = max(abs(od), abs(orr)) <= params$clock_max_accept_min &&
                          abs(orr - od) <= params$clock_max_accept_min,
               row.names = NULL)
  }))
  print(clock_table)
  cat(sprintf("\n%d of %d estimates accepted (offset and drift both within +/- %.0f min)\n",
              sum(clock_table$accepted), nrow(clock_table), params$clock_max_accept_min))
}
```

Even at 1-minute resolution the estimate is not free of doubt, so it is not
taken on trust. `hier_light` and `hier_light_forced` are identical except that
the second applies every estimate in defiance of the rule below, so section 7
measures what trusting this estimator costs rather than asserting that it must
be avoided.

The acceptance rule earned its keep the hard way. It used to reject only offsets
above 60 minutes against a 120-minute search, which caught estimates that had run
to the wall but was no test of plausibility at all. When the calibration of 5.2
improved, the profile likelihood started finding *interior* optima instead of
running to the bound, and those were duly accepted and applied: one tag was
handed 9 minutes at deployment ramping to 41.7 at recovery. `correct()` spreads
that disagreement linearly over the record, so it is about 8 degrees of longitude
by the end of the trip, and it cost the whole report several hundred kilometres
before the cause was found. The SPOT records printed above are what a clock of
this vintage actually does over a whole deployment: between -1.4 and +13.3
minutes. A 33-minute drift is two and a half times the largest real one, and an
estimate is now rejected if either its offset or its implied drift exceeds
`clock_max_accept_min`.

A better estimator is not the lesson here; a rejection rule that knows what a
clock can physically do is. `apply_clock_correction` is `FALSE` and no estimate
would pass the rule in any case, so the reported batches all run on raw
timestamps. `hier_light_forced` applies them anyway, and the contrast in section
7 is what the rule is worth: it is the difference between a threshold argued for
in prose and one with a number attached.

```{r clock-apply}
# Two copies. The panel itself is corrected only where an estimate passes the
# rule, and only if `apply_clock_correction` is set; with the rule as it stands
# nothing passes, so it keeps its raw timestamps.
#
# `tags_forced` applies every estimate REGARDLESS of the rule. That batch is not
# an alternative analysis, it is the evidence for the rule: without it the
# 20-minute threshold is a number asserted in prose, and with it section 7
# reports what trusting the estimator actually costs.
n_applied <- 0L; n_forced <- 0L
tags_forced <- tags
if (!is.null(clocks)) {
  for (id in names(tags)) {
    cl <- clocks[[id]]
    if (is.null(cl) || !is.function(cl$correct)) next
    tags_forced[[id]]$light$time <- cl$correct(tags_forced[[id]]$light$time)
    n_forced <- n_forced + 1L
    if (isTRUE(params$apply_clock_correction) &&
        isTRUE(clock_table$accepted[match(id, clock_table$id)])) {
      tags[[id]]$light$time <- cl$correct(tags[[id]]$light$time)
      n_applied <- n_applied + 1L
    }
  }
}
message("clock correction applied to ", n_applied, " of ", length(tags),
        " tags (forced onto ", n_forced, " for the contrast batch)")
```

## Study domain

```{r domain}
all_lon <- unlist(lapply(tags, function(g) g$argos$lon))
all_lat <- unlist(lapply(tags, function(g) g$argos$lat))
bbox <- c(xmin = floor(min(all_lon)) - 8, xmax = ceiling(max(all_lon)) + 4,
          ymin = floor(min(all_lat)) - 6, ymax = ceiling(max(all_lat)) + 8)
bbox[["ymin"]] <- max(bbox[["ymin"]], -89); bbox[["ymax"]] <- min(bbox[["ymax"]], 89)

date_rng  <- range(unlist(lapply(tags, function(g) c(g$t0, g$t1))))
all_dates <- seq(as.Date(as.POSIXct(date_rng[1], origin = "1970-01-01", tz = "UTC")) - 1,
                 as.Date(as.POSIXct(date_rng[2], origin = "1970-01-01", tz = "UTC")) + 1,
                 by = "day")
cat(sprintf("domain: lon %.0f..%.0f (0-360 E), lat %.0f..%.0f | %d days, %s to %s\n",
            bbox[["xmin"]], bbox[["xmax"]], bbox[["ymin"]], bbox[["ymax"]],
            length(all_dates), min(all_dates), max(all_dates)))
cat(sprintf("westernmost %.1f E, easternmost %.1f E; the colony is at %.2f E\n",
            min(all_lon), max(all_lon), tag_table$deploy_lon[1]))
```

# Environmental layers

Fetched once for the whole domain from ERDDAP and cached as GeoTIFFs. Both are
the native 0-360 variants, so nothing is reprojected or wrapped. `terra` objects
are external pointers and do not survive `saveRDS()`, which is why these caches
are rasters on disk, re-read on every knit.

```{r erddap-helpers}
erddap <- "https://coastwatch.pfeg.noaa.gov/erddap/griddap/"
enc <- function(s) gsub("]", "%5D", gsub("[", "%5B", s, fixed = TRUE), fixed = TRUE)

# Read a lon/lat(/time) ERDDAP NetCDF into a SpatRaster. ERDDAP latitude runs
# south to north; raster rows run north to south, so rows are reversed.
nc_to_rast <- function(path, var, has_time) {
  nc <- ncdf4::nc_open(path); on.exit(ncdf4::nc_close(nc))
  lon <- as.numeric(ncdf4::ncvar_get(nc, "longitude"))
  lat <- as.numeric(ncdf4::ncvar_get(nc, "latitude"))
  a   <- ncdf4::ncvar_get(nc, var, collapse_degen = TRUE)
  tm  <- if (has_time) as.numeric(ncdf4::ncvar_get(nc, "time")) else 0
  a   <- array(a, dim = c(length(lon), length(lat), length(tm)))
  ext <- terra::ext(min(lon) - abs(diff(lon)[1]) / 2, max(lon) + abs(diff(lon)[1]) / 2,
                    min(lat) - abs(diff(lat)[1]) / 2, max(lat) + abs(diff(lat)[1]) / 2)
  st <- terra::rast(lapply(seq_along(tm), function(i) {
    m <- t(a[, , i])
    terra::rast(m[rev(seq_len(nrow(m))), , drop = FALSE], extent = ext, crs = "EPSG:4326")
  }))
  if (has_time) names(st) <- format(as.Date(as.POSIXct(tm, origin = "1970-01-01", tz = "UTC")))
  st
}
```

## Sea surface temperature (NOAA OISST v2.1, daily, 0.25 degree)

```{r sst}
sst_tif <- file.path(params$cache_dir, "oisst.tif")
sst_dtf <- file.path(params$cache_dir, "oisst_dates.rds")

if (!file.exists(sst_tif) && isTRUE(params$fetch_env)) {
  chunks <- split(all_dates, ceiling(seq_along(all_dates) / 60))
  st <- lapply(seq_along(chunks), function(i) {
    ch <- chunks[[i]]
    message(sprintf("OISST %d/%d: %s..%s", i, length(chunks), min(ch), max(ch)))
    f <- file.path(tempdir(), sprintf("oisst_%02d.nc", i))
    u <- sprintf(paste0("%sncdcOisst21Agg.nc?sst[(%sT12:00:00Z):1:(%sT12:00:00Z)]",
                        "[(0.0):1:(0.0)][(%f):1:(%f)][(%f):1:(%f)]"),
                 erddap, min(ch), max(ch), bbox[["ymin"]], bbox[["ymax"]],
                 bbox[["xmin"]], bbox[["xmax"]])
    utils::download.file(enc(u), f, mode = "wb", quiet = TRUE)
    nc_to_rast(f, "sst", has_time = TRUE)
  })
  sst_stack <- do.call(c, st)
  terra::writeRaster(sst_stack, sst_tif, overwrite = TRUE)
  saveRDS(names(sst_stack), sst_dtf)
}

sst_by_date <- NULL
if (file.exists(sst_tif)) {
  sst_stack <- terra::rast(sst_tif)
  lyr_dates <- as.Date(readRDS(sst_dtf))
  names(sst_stack) <- format(lyr_dates)
  # build_aux_matrix() looks a layer up by "YYYY-MM-DD" and, when it is absent,
  # falls back to a network fetch needing rerddap. Guarantee a layer for every
  # date in range by mapping each to its nearest available layer, so that
  # fallback can never fire mid-fit.
  idx <- vapply(all_dates, function(d) which.min(abs(lyr_dates - d)), 1L)
  gap <- max(abs(as.numeric(lyr_dates[idx] - all_dates)))
  sst_by_date <- lapply(idx, function(i) sst_stack[[i]])
  names(sst_by_date) <- format(all_dates)
  cat(sprintf("OISST: %d layers, %d x %d cells, max |date gap| %d d\n",
              terra::nlyr(sst_stack), terra::ncol(sst_stack), terra::nrow(sst_stack), gap))
} else message("no SST layers; the SST term will be skipped")
```

## Bathymetry (ETOPO1, 0-360)

```{r bathy}
bathy_tif <- file.path(params$cache_dir, "bathy.tif")
if (!file.exists(bathy_tif) && isTRUE(params$fetch_env)) {
  f <- file.path(tempdir(), "etopo.nc")
  u <- sprintf("%setopo360.nc?altitude[(%f):%d:(%f)][(%f):%d:(%f)]", erddap,
               bbox[["ymin"]], params$bathy_stride, bbox[["ymax"]],
               bbox[["xmin"]], params$bathy_stride, bbox[["xmax"]])
  utils::download.file(enc(u), f, mode = "wb", quiet = TRUE)
  terra::writeRaster(nc_to_rast(f, "altitude", has_time = FALSE), bathy_tif, overwrite = TRUE)
}

depth_rast <- sea_rast <- NULL
if (file.exists(bathy_tif)) {
  elev <- terra::rast(bathy_tif)
  # bathy_source() and floor_rule() both want seafloor depth POSITIVE DOWN, with
  # land at 0 so a land cell cannot satisfy any non-trivial dive.
  depth_rast <- -1 * elev
  depth_rast[depth_rast < 0] <- 0
  names(depth_rast) <- "depth_m"
  sea_rast <- terra::classify(depth_rast, rbind(c(-Inf, 0.5, 0), c(0.5, Inf, 1)))
  names(sea_rast) <- "sea"
  cat(sprintf("ETOPO1: %d x %d cells, max depth %.0f m, %.1f%% sea\n",
              terra::ncol(depth_rast), terra::nrow(depth_rast),
              terra::global(depth_rast, "max", na.rm = TRUE)[1, 1],
              100 * terra::global(sea_rast, "mean", na.rm = TRUE)[1, 1]))
} else message("no bathymetry; depth and mask terms will be skipped")
```

## Is Mk9 temperature usable as an SST proxy? (and why SST is switched off)

The obvious second sensor is the external thermistor, matched against remotely
sensed SST. Whether that helps depends entirely on whether the reading is of the
water. Two things spoil it here. A hauled-out animal warms the sensor toward
body temperature. And an animal lying at the very surface has the sensor partly
in air. Both push the reading **up**, and because the North Pacific SST gradient
is meridional, a warm bias moves the estimate **south** by roughly 2.5 degrees of
latitude per degree Celsius.

Since the true position is known from Argos, this can be measured rather than
argued about: compare the tag reading with OISST at the animal's actual
position, stratified by how shallow it got.

```{r sst-validation}
sst_check <- if (is.null(sst_by_date)) NULL else do.call(rbind, lapply(names(tags), function(id) {
  g <- tags[[id]]
  tr <- interp_argos_xy(g$argos, g$light$time)
  d <- data.frame(id = id, depth_min = g$light$depth_min, depth_max = g$light$depth_max,
                  temp = g$light$temp_surf, lon = tr$lon, lat = tr$lat,
                  day = format(as.Date(g$light$time)))
  d <- d[is.finite(d$lon) & is.finite(d$temp), ]
  d$oisst <- NA_real_
  for (dy in unique(d$day)) {
    k <- which(d$day == dy)
    lyr <- sst_by_date[[dy]]
    if (!is.null(lyr)) d$oisst[k] <- as.numeric(terra::extract(lyr, cbind(d$lon[k], d$lat[k]))[, 1])
  }
  d$resid <- d$temp - d$oisst
  d[is.finite(d$resid), ]
}))

if (!is.null(sst_check)) {
  band <- cut(sst_check$depth_min, c(-Inf, 0.5, 2, 5, 10, Inf))
  print(data.frame(
    shallowest_point_m = levels(band),
    n = as.integer(table(band)),
    mean_resid_C = round(tapply(sst_check$resid, band, mean), 2),
    sd_C = round(tapply(sst_check$resid, band, stats::sd), 2),
    pct_over_2C = round(100 * tapply(sst_check$resid > 2, band, mean)),
    row.names = NULL))
  hauled <- sst_check$depth_max < 2          # never dived in the whole window
  cat(sprintf("\nwindows where the animal never dived (hauled out): n = %d, mean residual %+.2f C, max %+.1f C\n",
              sum(hauled), mean(sst_check$resid[hauled]), max(sst_check$resid[hauled])))
}
```

The thermistor is an **excellent** SST sensor between about 2 and 10 m, where it
is submerged but shallow, and an unusable one at the surface or hauled out. The
naive choice, taking the warmest near-surface reading in each knot, is the worst
possible reducer, because it selects precisely the contaminated readings.

Two consequences. First, if SST is used at all it must be restricted to the
2-10 m band with a low-order reducer, which is what `sst_depth_band` and the
`min` reducer in section 5.1 do. Second, and the reason `use_sst` is `FALSE` by
default here: after that filtering there are few enough qualifying readings, and
enough residual risk of contamination, that SST is not the term to test sensor
fusion with. **Depth is.** The tag's own depth channel against a bathymetric
chart is unambiguous, needs no calibration, and is the term that produces the
speed-up in section 8. Keeping SST off makes the fusion contrast a clean test of
one mechanism rather than a mixture of two, one of which is known to be biased.

```{r surf-series}
# The per-knot SST channel, if it is wanted: submerged but shallow only.
sst_band <- as.numeric(strsplit(params$sst_depth_band, ",")[[1]])
surf <- lapply(tags, function(g) {
  s <- g$light[is.finite(g$light$temp_surf) &
               g$light$depth_min >= sst_band[1] & g$light$depth_min <= sst_band[2],
               c("time", "temp_surf")]
  names(s)[2] <- "temp"
  s
})
data.frame(id = names(tags),
           n_windows = vapply(tags, function(g) nrow(g$light), 0L),
           n_in_band = vapply(surf, nrow, 0L),
           pct = round(100 * vapply(surf, nrow, 0L) /
                       vapply(tags, function(g) nrow(g$light), 0L), 1),
           row.names = NULL)
```

# Model configuration

## Sensor-fusion terms

Three terms, built per tag from that tag's own channels, in a soft and a hard
variant.

**Soft** (`floor_rule(sd = ...)`, `mask_rule(penalty = -20)`) tolerates
bathymetry-grid and depth-sensor error, but every cell still has to be
evaluated.

**Hard** (`floor_rule()`, `mask_rule(-Inf)`) declares a cell impossible, and the
engines then skip the light likelihood there entirely. That is where the
speed-up in section 8 comes from. To keep a hard constraint from excluding the
truth, the knot's deepest dive is reduced by `bathy_margin_m` first, so the
requirement is "the seabed is at least (deepest dive minus 300 m) deep" rather
than an exact match against a 7 km bathymetry grid.

```{r terms}
# build_aux_matrix() evaluates a source once per unique knot DATE, even for a
# static field. The closure below is rebuilt per tag and always sees that tag's
# own candidate coordinates, so memoising on first call is safe and removes a
# few hundred redundant raster extractions per tag.
memo_static <- function(fun) {
  cached <- NULL
  src <- function(lon, lat, date = NULL) {
    if (is.null(cached)) cached <<- fun(lon, lat, NULL)
    cached
  }
  class(src) <- c("tf_source", "function"); src
}

make_terms <- function(id, df = NULL, hard = FALSE) {
  g <- tags[[id]]; if (is.null(g)) return(list())
  out <- list()

  if (isTRUE(params$use_sst) && !is.null(sst_by_date) && nrow(surf[[id]]) >= 50) {
    out$sst <- location_term(
      name = "sst", tag = data.frame(time = surf[[id]]$time, value = surf[[id]]$temp),
      # Lowest reading in the knot, not the highest. Section 4.3 shows the only
      # contamination left after the depth-band filter is warm (residual air or
      # body heat), so a low-order reducer is the robust choice; taking the
      # warmest is the worst available option.
      reduce = "min",
      source = sst_source(rast = sst_by_date),
      rule   = student_rule(sd = params$sst_sd, df = params$sst_df))
  }

  if (!is.null(depth_rast)) {
    out$bathy <- location_term(
      name = "bathy", tag = data.frame(time = g$light$time, value = g$light$depth_max),
      reduce = if (hard) function(x) {
                 if (all(is.na(x))) NA_real_
                 else max(0, max(x, na.rm = TRUE) - params$bathy_margin_m)
               } else "max",
      source = memo_static(bathy_source(rast = depth_rast)),
      rule   = if (hard) floor_rule() else floor_rule(sd = params$bathy_sd))
  }

  if (!is.null(sea_rast)) {
    out$sea <- location_term(
      name = "sea", source = memo_static(raster_source(sea_rast)),
      rule = mask_rule(penalty = if (hard) -Inf else params$soft_mask_penalty))
  }
  unname(out)
}
soft_terms <- function(id, df) make_terms(id, df, hard = FALSE)
hard_terms <- function(id, df) make_terms(id, df, hard = TRUE)

term_inventory <- do.call(rbind, lapply(names(tags), function(id) {
  tm <- make_terms(id)
  data.frame(id = id, n_terms = length(tm),
             terms = paste(vapply(tm, function(t) t$name, ""), collapse = "+"),
             row.names = NULL)
}))
term_inventory
```

## The tag's light response

The engines model clear-sky light as a clamped linear function of solar zenith
angle. Where that line sits is a property of the **light channel**, and it is not
the same for every tag. A logger whose output is linear in irradiance collapses
within civil twilight, over about 11 degrees of zenith. The Mk9 reports a
compressed, roughly logarithmic light level whose clear-sky curve declines
smoothly over about 25 degrees and settles on a non-zero dark reading. Latitude
is read from that gradient, so assuming the wrong one is a first-order error,
and archival tags differ in it even within a batch. It is therefore measured per
tag rather than assumed.

`fit_light_response()` does this at the release site, where the position is
known, by fitting the **upper envelope** of light against zenith. Shading, cloud
and depth can only ever reduce the measured light, so the clear-sky curve is the
top of the point cloud rather than its middle, which is what lets the fit
survive an animal that is underwater most of the time.

Two details of *which* record to fit turn out to matter more than anything else
in this report, so they are worth stating before the code.

**The window has to be one where the animal really is where you say it is.**
Fitting a fixed fifteen days from the release fix assumes the animal is still at
the colony, and these animals are not: by the end of that window the Argos track
puts one of them a median 450 km away. A moving animal smears the envelope
across zenith and flattens the fitted curve, and latitude is read from precisely
that gradient. Tag 2021025 came out with a 54 degree transition against the 30
degrees it shows when fitted at the positions it actually occupied, and it was
the worst track in the panel at 1255 km. The tag dates its own departure without
any help from Argos: a hauled-out seal does not dive, so the last day before
continuous diving begins is the last day the colony position is true.

**The geometry and the intensity scale want different windows.** `z50` and the
transition width need known positions, so a haul-out. `baseline` and `max_light`
need a record whose *shading* looks like the record being tracked, and a
hauled-out seal never dives, so its fifth percentile is night at the surface
rather than a deep daytime dive. Taking both from the haul-out is worse than the
naive recipe (447 km against 321 km on a three-tag trial); taking only the
geometry from it, and the scale from the record itself, gave 104 km. That split
is what `scale_light` exists for.

One consequence is convenient. The geometry is a property of the light channel,
not the animal, and all ten seals carry the same tag model, so the seven usable
haul-out fits are seven noisy measurements of one curve. Pooling them with
`pool_light_responses()` improves even the animals that contributed a good fit
of their own, and it is the only way to give a response to the three animals
that left within a day of being tagged and can never be calibrated. The
comparison behind these claims is run, not asserted: see *Choosing the
calibration window* below.

```{r light-response}
# Everything here is fitted on the working (30-minute) series, not on the
# high-resolution endpoint slice: the response is a shape in zenith, not an edge
# in time, so it wants many days at coarse resolution rather than a few at fine.
# Fitted on three days, three of these ten tags returned physically impossible
# curves (transitions of 88 to 107 degrees, saturating only above zenith 30).
#
# When did the animal leave? A hauled-out seal does not dive, so take the first
# day after which it is never ashore again. This uses the depth channel only:
# no Argos, no assumption beyond the release fix.
departure_time <- function(d, dive_m = 50) {
  b <- as.integer(floor(as.numeric(d$time) / 86400))
  deep <- tapply(d$depth_max > dive_m, b, mean, na.rm = TRUE)
  day <- as.integer(names(deep))
  at_sea <- deep > 0.5
  if (!any(at_sea)) return(NA_real_)
  last_ashore <- if (any(!at_sea)) max(which(!at_sea)) else 0L
  i <- if (last_ashore >= length(at_sea)) which(at_sea)[1] else last_ashore + 1L
  day[i] * 86400
}

# The intensity scale, per tag, from the standard window: this one has to look
# like the record being tracked, dives and all.
scale_fits <- lapply(tags, function(g) {
  k <- g$light$time <= min(g$light$time) + params$cal_window_days * 86400
  fit_light_response(g$light$time[k], g$light$light[k],
                     lon = g$deploy[["lon"]], lat = g$deploy[["lat"]])
})

# The geometry, per tag, from the pre-departure haul-out: this one has to be a
# window where the release fix is still the truth. Animals that left too fast to
# leave 200 usable samples contribute nothing here and rely on the pool.
geom_fits <- lapply(tags, function(g) {
  dep <- departure_time(g$light)
  k <- is.finite(dep) & as.numeric(g$light$time) < dep
  if (sum(k) < 200) return(NULL)
  # No `scale_light` here: only z50 and the transition width survive pooling,
  # and neither depends on it. The scale comes from `scale_fits`.
  fit_light_response(g$light$time[k], g$light$light[k],
                     lon = g$deploy[["lon"]], lat = g$deploy[["lat"]])
})

haulout_table <- do.call(rbind, lapply(names(tags), function(id) {
  dep <- departure_time(tags[[id]]$light)
  k <- is.finite(dep) & as.numeric(tags[[id]]$light$time) < dep
  data.frame(id = id,
             departs = if (is.finite(dep))
               format(as.POSIXct(dep, origin = "1970-01-01", tz = "UTC"), "%Y-%m-%d")
               else NA_character_,
             haulout_days = round(sum(k) * params$decimate_minutes / 1440, 1),
             n = sum(k), used = !is.null(geom_fits[[id]]), row.names = NULL)
}))
haulout_table

# Pool the geometry over the tags that have one; every tag keeps its own scale.
responses <- pool_light_responses(geom_fits, scale_fits)
pooled_geom <- attr(responses, "pooled")
cat(sprintf("\npooled geometry from %d of %d tags: z50 = %.2f, transition %.1f degrees\n",
            pooled_geom[["n"]], length(tags), pooled_geom[["z50"]],
            4.394 * pooled_geom[["scale"]]))
cat("per-tag haul-out geometry, before pooling:\n")
print(data.frame(
  id    = names(geom_fits)[!vapply(geom_fits, is.null, logical(1))],
  z50   = round(vapply(Filter(Negate(is.null), geom_fits), function(r) r$z50, 0), 2),
  width = round(vapply(Filter(Negate(is.null), geom_fits), function(r) r$width_deg, 0), 1),
  row.names = NULL))

# Pooling has replaced the old rescue rule, which substituted a panel median
# only for tags whose fitted transition was physically impossible (outside 5 to
# 60 degrees). That rule caught the catastrophic failures and missed the merely
# bad ones: tag 2021025's 54 degree transition passed it, and that tag was the
# worst track in the panel. Pooling every tag rather than only the obviously
# broken ones removes the need for a threshold at all. The check is kept as an
# assertion, because a pooled geometry outside this range would mean the whole
# panel had failed and nothing downstream would be worth reading.
stopifnot(pooled_geom[["z50"]] > 80, pooled_geom[["z50"]] < 105,
          4.394 * pooled_geom[["scale"]] > 5, 4.394 * pooled_geom[["scale"]] < 60)

resp_table <- do.call(rbind, lapply(names(responses), function(id) {
  r <- responses[[id]]
  if (is.null(r)) return(data.frame(id = id, slope = NA_real_, width_deg = NA_real_,
                                    saturate_at = NA_real_, zero_at = NA_real_,
                                    dark_level = NA_real_, baseline = NA_real_,
                                    own_geometry = NA))
  data.frame(id = id, slope = round(r$slope, 2), width_deg = round(r$width_deg, 1),
             saturate_at = round(r$saturate_at, 1), zero_at = round(r$zero_at, 1),
             dark_level = if (is.finite(r$dark_level)) round(r$dark_level) else NA_real_,
             baseline = round(r$baseline),
             own_geometry = !is.null(geom_fits[[id]]), row.names = NULL)
}))
resp_table
cat(sprintf("\nmedian fitted slope %.2f over a %.0f degree transition\n",
            stats::median(resp_table$slope, na.rm = TRUE),
            stats::median(resp_table$width_deg, na.rm = TRUE)))
cat(sprintf("the old hard-coded default would have been range/11 = %.1f over 11 degrees\n",
            stats::median(vapply(tags, function(g)
              as.numeric(diff(stats::quantile(g$light$light, c(.05, .95)))) / 11, 0))))
```

Three things to read off this table. `width_deg` well above 11 says the channel
is log-scaled. It is now identical across tags, because the geometry is pooled;
the per-tag spread before pooling is printed above it, and its tightness is the
evidence that pooling is reasonable. `dark_level` sits far above `baseline` (the
5th percentile the engines subtract) because for a continuously diving animal the
5th percentile is a **deep daytime dive**, which is darker than night at the
surface. The dark level is reported as a diagnostic but deliberately not used as
the baseline: subtracting it clips every deep observation to zero and destroys
the gradient that carries the signal, and doing so is worth about 300 km of
median error.

```{r response-plot, fig.height = 4.5}
op <- par(mar = c(4.2, 4.2, 2.4, 0.8))
zz <- seq(40, 130, by = 0.5)
ymax <- max(vapply(responses, function(r)
  if (is.null(r)) 0 else r$baseline + r$max_light, 0)) * 1.15
plot(NA, xlim = c(40, 130), ylim = c(0, ymax),
     xlab = "solar zenith angle (degrees)", ylab = "light (tag units)",
     main = "Fitted clear-sky response, all tags")
# The envelopes drawn are the haul-out ones the geometry was actually measured
# from, not the 15-day ones that only set the intensity scale. Drawing the
# latter under a pooled curve would invite the reader to check the fit against
# data it was not fitted to.
for (id in names(geom_fits)) {
  r <- geom_fits[[id]]; if (is.null(r) || is.null(r$envelope)) next
  lines(r$envelope$zenith, r$envelope$light, col = "#95a5a655")
}
for (id in names(responses)) {
  r <- responses[[id]]; if (is.null(r)) next
  lines(zz, pmin(pmax(r$calibration[1] - r$calibration[2] * zz, 0), r$max_light) + r$baseline,
        col = "#c0392b", lwd = 1.2)
}
g1 <- tags[[1]]
rng1 <- as.numeric(diff(stats::quantile(g1$light$light, c(.05, .95))))
q051 <- as.numeric(stats::quantile(g1$light$light, .05))
lines(zz, pmin(pmax(rng1 / 11 * 96 - rng1 / 11 * zz, 0), rng1) + q051,
      col = "#2980b9", lwd = 2, lty = 2)
legend("bottomleft", c("haul-out envelope", "pooled response", "old hard-coded default"),
       col = c("#95a5a6", "#c0392b", "#2980b9"), lwd = c(1, 1.2, 2),
       lty = c(1, 1, 2), bty = "n", cex = 0.9)
par(op)
```

## The surrogate mesh

`TwilightFreeHier()` proposes from a coarse grid-HMM smoother on an internal
mesh spanning `range(deploy, retrieve) +/- mesh_pad`. Deployment and recovery
are the same colony here, so that box is a single point and the padding sets the
mesh outright. These animals range about 60 degrees west but only about 20
degrees north, and a single scalar pad wide enough for the longitude would put
mesh rows past the pole, where `solar_zenith()` silently returns mirror-point
geometry and those cells pollute the proposal's mean and covariance. `mesh_pad`
therefore takes `c(lon_pad, lat_pad)`, and the latitude range is clamped to
+/- 88 degrees.

```{r mesh}
pad  <- c(params$mesh_pad_lon, params$mesh_pad_lat)
lon0 <- range(c(tag_table$deploy_lon, tag_table$recover_lon))
lat0 <- range(c(tag_table$deploy_lat, tag_table$recover_lat))
need_lon <- max(lon0[1] - min(all_lon), max(all_lon) - lon0[2]) + 4
need_lat <- max(lat0[1] - min(all_lat), max(all_lat) - lat0[2]) + 4
cat(sprintf("mesh_pad = c(%.0f, %.0f) | needed: lon >= %.0f, lat >= %.0f\n",
            pad[1], pad[2], need_lon, need_lat))
if (pad[1] < need_lon) warning("longitude pad is smaller than the observed range")
if (pad[2] < need_lat) warning("latitude pad is smaller than the observed range")
mesh_lon <- seq(lon0[1] - pad[1], lon0[2] + pad[1], by = params$coarse_res)
mesh_lat <- seq(max(-88, lat0[1] - pad[2]), min(88, lat0[2] + pad[2]), by = params$coarse_res)
cat(sprintf("surrogate mesh: lon %.0f..%.0f, lat %.0f..%.0f, %d x %d = %d cells\n",
            min(mesh_lon), max(mesh_lon), min(mesh_lat), max(mesh_lat),
            length(mesh_lon), length(mesh_lat), length(mesh_lon) * length(mesh_lat)))
```

`coarse_res` also sets the resolution at which this engine sees the auxiliary
terms: the SST field a knot sees is piecewise constant on 2-degree cells. In the
North Pacific transition zone SST falls by roughly 0.5-1 degree C per degree of
latitude, so a 2-degree mesh quantises the SST constraint into steps comparable
to `sst_sd`. That is the main approximation in the hierarchical fusion path, and
it is what the grid HMM batch, on 1-degree cells, is there to check.

## The HMM grid

```{r grid}
grid <- terra::rast(xmin = floor(bbox[["xmin"]]), xmax = ceiling(bbox[["xmax"]]),
                    ymin = floor(bbox[["ymin"]]), ymax = ceiling(bbox[["ymax"]]),
                    resolution = params$grid_cell_deg, crs = "EPSG:4326")
terra::values(grid) <- 1
cat(sprintf("grid HMM: %d x %d = %d cells at %.1f degrees\n",
            terra::ncol(grid), terra::nrow(grid), terra::ncell(grid), params$grid_cell_deg))
```

## Panel inputs

```{r panel}
# Every engine gets that tag's fitted response, so none falls back to the
# hard-coded one. Light is baselined with the fitted baseline (the 5th
# percentile), not the fitted dark level: see 5.2.
tag_light <- function(g) {
  r <- responses[[g$id]]
  base <- if (is.null(r)) as.numeric(stats::quantile(g$light$light, .05)) else r$baseline
  pmax(0, g$light$light - base)
}
tag_cal <- function(g) { r <- responses[[g$id]]; if (is.null(r)) NULL else r$calibration }
tag_lp  <- function(g) {
  r <- responses[[g$id]]
  if (is.null(r)) NULL else c(1 / (r$max_light * 0.5), r$max_light, 0.10)
}

panel <- function(tg) list(
  data = stats::setNames(lapply(tg, function(g)
    data.frame(Date = g$light$time, Light = tag_light(g))), names(tg)),
  locations = data.frame(
    id = names(tg),
    deploy_lon   = vapply(tg, function(g) unname(g$deploy[["lon"]]), 0),
    deploy_lat   = vapply(tg, function(g) unname(g$deploy[["lat"]]), 0),
    retrieve_lon = vapply(tg, function(g) unname(g$recover[["lon"]]), 0),
    retrieve_lat = vapply(tg, function(g) unname(g$recover[["lat"]]), 0), row.names = NULL))
P     <- panel(tags)
P_forced <- panel(tags_forced)
P$locations
```

# Fitting

## Hierarchical batches

```{r hier}
# TwilightFreeHier applies one calibration across the panel, so the per-tag
# responses are pooled by their median. Since 5.2 now pools the geometry
# already, this only pools what is left, the per-tag intensity scale.
#
# Worth noticing in passing: before 5.2 pooled anything, this line was the only
# pooling in the report, and the hierarchical batch was also the only one that
# performed well (224 km against the grid engine's 366 km on the same tags). The
# hierarchical model was not beating the grid HMM on its merits so much as being
# handed a calibration that had been quietly averaged over ten tags. Pooling the
# geometry properly, for every engine, is what closes that gap.
cal_panel <- {
  ok <- Filter(Negate(is.null), responses)
  c(stats::median(vapply(ok, function(r) r$calibration[1], 0)),
    stats::median(vapply(ok, function(r) r$calibration[2], 0)))
}
lp_panel <- {
  ok <- Filter(Negate(is.null), responses)
  m <- stats::median(vapply(ok, function(r) r$max_light, 0))
  c(1 / (m * 0.5), m, 0.10)
}

fit_hier <- function(pp, terms) timed(TwilightFreeHier(
  data = pp$data, locations = pp$locations, terms = terms,
  calibration = cal_panel, likelihood_params = lp_panel,
  step_hours = params$step_hours, coarse_res = params$coarse_res, mesh_pad = pad,
  surrogate_diffusion = params$surrogate_diffusion,
  sweeps = as.integer(params$sweeps), burn = as.integer(params$burn),
  thin = as.integer(params$thin), metric = "spherical", seed = params$seed))

hier_light     <- cache_rds("fit_hier_light_v2",     fit_hier(P,     NULL))
hier_fusion    <- cache_rds("fit_hier_fusion_v2",    fit_hier(P,     soft_terms))
hier_light_forced <- cache_rds("fit_hier_light_forced_v2", fit_hier(P_forced, NULL))

for (nm in c("hier_light", "hier_fusion", "hier_light_forced"))
  cat(sprintf("%-16s %7.1f s (%.1f min)\n", nm, get(nm)$seconds, get(nm)$seconds / 60))
print(hier_light$value)
print(hier_fusion$value)
```

## Grid HMM batches

This is the engine in which a hard constraint pays for itself, because candidate
locations are enumerated explicitly: a cell excluded by bathymetry or by the
coastline never has its light likelihood evaluated at all.

```{r grid-fits}
fit_grid_one <- function(id, terms) {
  g <- tags[[id]]
  invisible(capture.output(
    f <- TwilightFreeGrid(g$light$time, tag_light(g), grid = grid,
                          start_lon = unname(g$deploy[["lon"]]),
                          start_lat = unname(g$deploy[["lat"]]),
                          end_lon   = unname(g$recover[["lon"]]),
                          end_lat   = unname(g$recover[["lat"]]),
                          step_hours = params$step_hours,
                          diffusion = params$grid_diffusion, terms = terms,
                          calibration = tag_cal(g), likelihood_params = tag_lp(g))))
  # Keep only what the comparison needs: the full posterior is K x n doubles.
  list(time = f$fit$time, lon = f$fit$lon, lat = f$fit$lat, log_z = f$log_z)
}
fit_grid_batch <- function(terms_fn) timed(stats::setNames(lapply(names(tags), function(id) {
  message("grid: ", id)
  fit_grid_one(id, if (is.null(terms_fn)) list() else terms_fn(id, NULL))
}), names(tags)))

grid_light  <- cache_rds("fit_grid_light_v2",  fit_grid_batch(NULL))
grid_fusion <- cache_rds("fit_grid_fusion_v2", fit_grid_batch(hard_terms))
cat(sprintf("grid_light  %7.1f s (%.1f min)\ngrid_fusion %7.1f s (%.1f min)\n",
            grid_light$seconds, grid_light$seconds / 60,
            grid_fusion$seconds, grid_fusion$seconds / 60))
```

## FFBS on a subset

The hierarchical engine proposes from a coarse mesh and the grid HMM discretises
space; FFBS works in continuous space with neither, so it checks whether
discretisation is costing accuracy. It is `O(N^2)` per knot and single-track, so
it runs on the tags with the densest Argos coverage, where the comparison can be
made most precisely.

```{r ffbs}
ffbs_ids <- names(sort(vapply(tags, function(g) nrow(g$argos) / max(g$days, 1), 0),
                       decreasing = TRUE))[seq_len(min(params$ffbs_n, length(tags)))]
cat("FFBS subset:", paste(ffbs_ids, collapse = ", "), "\n")

# The SMC engine rasterises auxiliary terms onto a background grid, and when no
# `spatial_mask` is supplied that grid defaults to 2 degrees spanning -180..180.
# This analysis works in 0..360, so every particle would fall outside that
# extent, be clamped to the edge column, and receive a constant auxiliary value:
# the terms would silently do nothing. Supplying a mask over the study domain
# fixes the convention. It is all-ones inside the domain, so it is the same mild
# "stay in the study box" constraint in BOTH ffbs batches and cannot confound the
# light-versus-fusion contrast.
smc_domain <- terra::rast(xmin = floor(bbox[["xmin"]]), xmax = ceiling(bbox[["xmax"]]),
                          ymin = floor(bbox[["ymin"]]), ymax = ceiling(bbox[["ymax"]]),
                          resolution = 1, crs = "EPSG:4326")
terra::values(smc_domain) <- 1

run_ffbs <- function(id, terms) {
  g <- tags[[id]]
  TwilightFreeSMC(date_time = g$light$time, light = tag_light(g),
    calibration = tag_cal(g), likelihood_params = tag_lp(g),
    n_particles = as.integer(params$ffbs_particles),
    start_lon = unname(g$deploy[["lon"]]),  start_lat = unname(g$deploy[["lat"]]),
    end_lon   = unname(g$recover[["lon"]]), end_lat   = unname(g$recover[["lat"]]),
    method = "ffbs", step_hours = params$step_hours,
    diffusion = params$grid_diffusion, spatial_mask = smc_domain,
    terms = terms, seed = params$seed)
}
ffbs_light  <- cache_rds("fit_ffbs_light_v2", timed(stats::setNames(lapply(ffbs_ids, function(id) {
  message("ffbs light: ", id); run_ffbs(id, list()) }), ffbs_ids)))
ffbs_fusion <- cache_rds("fit_ffbs_fusion_v2", timed(stats::setNames(lapply(ffbs_ids, function(id) {
  message("ffbs fusion: ", id); run_ffbs(id, soft_terms(id, NULL)) }), ffbs_ids)))
cat(sprintf("ffbs_light  %7.1f s\nffbs_fusion %7.1f s\n",
            ffbs_light$seconds, ffbs_fusion$seconds))
```

# Validation against Argos

Argos positions are interpolated to the fitted knot times, but only where a knot
is bracketed by fixes no more than `argos_max_gap_h` apart; other knots are
scored as missing rather than compared with a guess. Argos error for these
classes is of order 1-10 km against a light-geolocation error of order 100 km,
so treating it as truth is safe, and the interpolation, not the Argos error, is
what the gap rule guards against.

```{r scoring}
in_equinox <- function(times, half_width = 21) {
  yrs <- unique(as.integer(format(times, "%Y")))
  eq <- as.POSIXct(c(paste0(yrs, "-03-20"), paste0(yrs, "-09-22")), tz = "UTC")
  apply(outer(as.numeric(times), as.numeric(eq),
              function(a, b) abs(a - b) / 86400), 1, min) <= half_width
}

score_tag <- function(id, knot_time, est_lon, est_lat, lon_sd, lat_sd, engine, batch) {
  est_lon <- lon360(est_lon)
  truth <- interp_argos_xy(tags[[id]]$argos, knot_time, params$argos_max_gap_h)
  data.frame(id = id, engine = engine, batch = batch, time = knot_time,
             est_lon = est_lon, est_lat = est_lat,
             true_lon = truth$lon, true_lat = truth$lat,
             err_km = gc_dist_km(est_lon, est_lat, truth$lon, truth$lat),
             err_lon = dlon(est_lon, truth$lon), err_lat = est_lat - truth$lat,
             lon_sd = if (is.null(lon_sd)) NA_real_ else lon_sd,
             lat_sd = if (is.null(lat_sd)) NA_real_ else lat_sd,
             equinox = in_equinox(knot_time), row.names = NULL)
}

summarise_tag <- function(k) {
  kk <- k[is.finite(k$err_km), , drop = FALSE]; eq <- kk$equinox
  data.frame(id = k$id[1], engine = k$engine[1], batch = k$batch[1],
    n_knots = nrow(k), n_scored = nrow(kk),
    median_km = stats::median(kk$err_km), rmse_km = sqrt(mean(kk$err_km^2)),
    q90_km = unname(stats::quantile(kk$err_km, .9)),
    bias_lon = mean(kk$err_lon), bias_lat = mean(kk$err_lat),
    rmse_lon_deg = sqrt(mean(kk$err_lon^2)), rmse_lat_deg = sqrt(mean(kk$err_lat^2)),
    median_km_equinox = if (any(eq))  stats::median(kk$err_km[eq])  else NA_real_,
    median_km_other   = if (any(!eq)) stats::median(kk$err_km[!eq]) else NA_real_,
    cover95_lon = if (all(is.na(kk$lon_sd))) NA_real_ else
                    mean(abs(kk$err_lon) <= 1.96 * kk$lon_sd),
    cover95_lat = if (all(is.na(kk$lat_sd))) NA_real_ else
                    mean(abs(kk$err_lat) <= 1.96 * kk$lat_sd), row.names = NULL)
}
```

```{r score-all}
score_hier <- function(fit, label) do.call(rbind, lapply(names(fit$tracks), function(id) {
  tr <- fit$tracks[[id]]
  score_tag(id, tr$time, tr$lon, tr$lat, tr$lon_sd, tr$lat_sd, "hier", label) }))
score_grid <- function(fits, label) do.call(rbind, lapply(names(fits), function(id) {
  f <- fits[[id]]
  score_tag(id, as.POSIXct(f$time, origin = "1970-01-01", tz = "UTC"),
            f$lon, f$lat, NULL, NULL, "grid", label) }))
score_smc <- function(fits, label) do.call(rbind, lapply(names(fits), function(id) {
  r <- fits[[id]]
  score_tag(id, as.POSIXct(r$knot_times, origin = "1970-01-01", tz = "UTC"),
            r$lon, r$lat, r$lon_sd, r$lat_sd, "ffbs", label) }))

knots <- rbind(
  score_hier(hier_light$value,     "light"),
  score_hier(hier_fusion$value,    "light+env"),
  score_hier(hier_light_forced$value, "light, forced clock"),
  score_grid(grid_light$value,     "light"),
  score_grid(grid_fusion$value,    "light+env"),
  score_smc(ffbs_light$value,      "light"),
  score_smc(ffbs_fusion$value,     "light+env"))

results <- do.call(rbind, lapply(split(knots, list(knots$id, knots$engine, knots$batch),
                                       drop = TRUE), summarise_tag))
row.names(results) <- NULL
write.csv(results, file.path(params$out_dir, "per_tag_results.csv"), row.names = FALSE)
write.csv(knots,   file.path(params$out_dir, "per_knot_errors.csv"), row.names = FALSE)
results[order(results$engine, results$batch, results$id),
        c("id", "engine", "batch", "n_scored", "median_km", "rmse_km",
          "bias_lon", "bias_lat", "cover95_lon", "cover95_lat")]
```

## Choosing the calibration window

Section 5.2 asserted that where the response is fitted matters more than
anything else in this report. This runs the comparison rather than asserting it,
on the grid HMM, over every tag's whole deployment, changing nothing but the
response handed to the engine.

`naive` is the recipe this report used until it was checked: fit each tag over a
fixed 15 days from the release fix, assuming the animal stays put, and use that
fit whole. `pooled` is the recipe above: geometry from each animal's
pre-departure haul-out, pooled across the panel, with each animal keeping its own
intensity scale. Nothing else differs, and neither recipe uses an Argos position
for anything except scoring afterwards.

```{r calibration-window}
naive_responses <- stats::setNames(lapply(tags, function(g) {
  k <- g$light$time <= min(g$light$time) + params$cal_window_days * 86400
  fit_light_response(g$light$time[k], g$light$light[k],
                     lon = g$deploy[["lon"]], lat = g$deploy[["lat"]])
}), names(tags))

fit_grid_with <- function(resp) timed(stats::setNames(lapply(names(tags), function(id) {
  g <- tags[[id]]; r <- resp[[id]]
  if (is.null(r)) return(NULL)
  message("calibration check: ", id)
  invisible(capture.output(
    f <- TwilightFreeGrid(g$light$time, pmax(0, g$light$light - r$baseline), grid = grid,
                          start_lon = unname(g$deploy[["lon"]]),
                          start_lat = unname(g$deploy[["lat"]]),
                          end_lon   = unname(g$recover[["lon"]]),
                          end_lat   = unname(g$recover[["lat"]]),
                          step_hours = params$step_hours,
                          diffusion = params$grid_diffusion,
                          calibration = r$calibration,
                          likelihood_params = c(1/(r$max_light*0.5), r$max_light, 0.10))))
  list(time = f$fit$time, lon = f$fit$lon, lat = f$fit$lat)
}), names(tags)))

cal_naive  <- cache_rds("cal_check_naive",  fit_grid_with(naive_responses))
cal_pooled <- cache_rds("cal_check_pooled", fit_grid_with(responses))

cal_knots <- rbind(score_grid(Filter(Negate(is.null), cal_naive$value),  "naive"),
                   score_grid(Filter(Negate(is.null), cal_pooled$value), "pooled"))
cal_res <- do.call(rbind, lapply(split(cal_knots, list(cal_knots$id, cal_knots$batch),
                                       drop = TRUE), summarise_tag))
row.names(cal_res) <- NULL

# Per tag, side by side, so a single rescued animal cannot carry the average.
cal_wide <- merge(
  cal_res[cal_res$batch == "naive",  c("id", "median_km", "rmse_lat_deg", "bias_lat")],
  cal_res[cal_res$batch == "pooled", c("id", "median_km", "rmse_lat_deg", "bias_lat")],
  by = "id", suffixes = c("_naive", "_pooled"))
cal_wide[] <- lapply(cal_wide, function(x) if (is.numeric(x)) round(x, 2) else x)
cal_wide$own_geometry <- !vapply(geom_fits[cal_wide$id], is.null, logical(1))
cal_wide

cat("\npooled over knots:\n")
print(do.call(rbind, lapply(split(cal_knots, cal_knots$batch), function(k) {
  kk <- k[is.finite(k$err_km), , drop = FALSE]
  data.frame(recipe = k$batch[1], n_scored = nrow(kk),
             median_km = round(stats::median(kk$err_km)),
             q90_km = round(unname(stats::quantile(kk$err_km, .9))),
             bias_lat = round(mean(kk$err_lat), 2),
             rmse_lat = round(sqrt(mean(kk$err_lat^2)), 2),
             row.names = NULL)
})), row.names = FALSE)
```

The three animals with no haul-out are worth reading separately: they contribute
no geometry to the pool and take it entirely on loan, so they are the honest test
of whether a response measured on other individuals transfers. If they improve,
the geometry really is a property of the tag model rather than of the animal.

## Batch summary

```{r batch-summary}
# Pooled over knots, not averaged over per-tag summaries, so a 248-day
# deployment carries more weight than a 137-day one. Coverage is averaged over
# tags, because it is a property of the reported intervals rather than of a knot.
agg <- function(k, r) {
  kk <- k[is.finite(k$err_km), , drop = FALSE]; eq <- kk$equinox
  data.frame(n_tags = length(unique(kk$id)), n_knots = nrow(kk),
    median_km = round(stats::median(kk$err_km), 1),
    rmse_km   = round(sqrt(mean(kk$err_km^2)), 1),
    q90_km    = round(unname(stats::quantile(kk$err_km, .9)), 1),
    rmse_lon_deg = round(sqrt(mean(kk$err_lon^2)), 2),
    rmse_lat_deg = round(sqrt(mean(kk$err_lat^2)), 2),
    equinox_km = if (any(eq))  round(stats::median(kk$err_km[eq]), 1)  else NA_real_,
    other_km   = if (any(!eq)) round(stats::median(kk$err_km[!eq]), 1) else NA_real_,
    cover95_lon = round(mean(r$cover95_lon, na.rm = TRUE), 2),
    cover95_lat = round(mean(r$cover95_lat, na.rm = TRUE), 2))
}
keys <- unique(paste(results$engine, results$batch, sep = " / "))
batch <- do.call(rbind, lapply(keys, function(kk) {
  p <- strsplit(kk, " / ", fixed = TRUE)[[1]]
  cbind(batch = kk, agg(knots[knots$engine == p[1] & knots$batch == p[2], ],
                        results[results$engine == p[1] & results$batch == p[2], ]))
}))
row.names(batch) <- NULL
batch
```

## Paired contrasts

```{r paired}
paired <- function(engine, a, b, label) {
  m <- merge(results[results$engine == engine & results$batch == a, ],
             results[results$engine == engine & results$batch == b, ],
             by = "id", suffixes = c("_a", "_b"))
  if (!nrow(m)) return(NULL)
  data.frame(contrast = label, engine = engine, n = nrow(m),
    d_median_km = round(mean(m$median_km_b - m$median_km_a), 1),
    pct_change  = round(100 * mean((m$median_km_b - m$median_km_a) / m$median_km_a), 1),
    improved    = sum(m$median_km_b < m$median_km_a),
    d_lat_deg   = round(mean(m$rmse_lat_deg_b - m$rmse_lat_deg_a), 2),
    d_lon_deg   = round(mean(m$rmse_lon_deg_b - m$rmse_lon_deg_a), 2),
    d_equinox_km = round(mean(m$median_km_equinox_b - m$median_km_equinox_a, na.rm = TRUE), 1),
    p_wilcoxon  = tryCatch(signif(stats::wilcox.test(m$median_km_a, m$median_km_b,
                                                     paired = TRUE)$p.value, 3),
                           error = function(e) NA_real_), row.names = NULL)
}
contrasts <- do.call(rbind, Filter(Negate(is.null), list(
  paired("hier", "light", "light+env", "sensor fusion (soft)"),
  paired("grid", "light", "light+env", "sensor fusion (hard)"),
  paired("ffbs", "light", "light+env", "sensor fusion (soft)"),
  paired("hier", "light", "light, forced clock", "applying the estimated clock"))))
contrasts
```

Negative `d_median_km` means the second batch was better. `d_lat_deg` versus
`d_lon_deg` separates the two mechanisms: SST in the North Pacific varies far
more strongly with latitude than with longitude, so an SST term that is working
shows up as a drop in latitude error, concentrated in the equinox window, while
leaving longitude alone. A clock shift, conversely, moves longitude and nothing
else, so `applying the estimated clock` should show up almost entirely in
`d_lon_deg`. It does, which is the signature that identified it: a timing error
cannot masquerade as anything but a longitude error, and no other candidate
explanation for the discrepancy has that shape.

# Timing, and whether fusion pays for itself

```{r timing}
tag_days  <- sum(vapply(tags, function(g) g$days, 0))
ffbs_days <- sum(vapply(tags[ffbs_ids], function(g) g$days, 0))
timing <- data.frame(
  batch  = c("hier_light", "hier_fusion", "hier_light_forced",
             "grid_light", "grid_fusion", "ffbs_light", "ffbs_fusion"),
  engine = c(rep("TwilightFreeHier", 3), rep("TwilightFreeGrid", 2),
             rep("TwilightFreeSMC(ffbs)", 2)),
  terms  = c("none", "soft bathy+SST+mask", "none",
             "none", "HARD bathy+mask+SST", "none", "soft bathy+SST"),
  n_tags = c(rep(length(tags), 5), rep(length(ffbs_ids), 2)),
  seconds = round(c(hier_light$seconds, hier_fusion$seconds, hier_light_forced$seconds,
                    grid_light$seconds, grid_fusion$seconds,
                    ffbs_light$seconds, ffbs_fusion$seconds), 1))
timing$minutes <- round(timing$seconds / 60, 2)
timing$sec_per_tag_day <- round(timing$seconds /
  ifelse(grepl("^ffbs", timing$batch), ffbs_days, tag_days), 3)
timing

cat(sprintf("\nGrid HMM, HARD constraints vs light only    : %+.1f%% wall clock\n",
            100 * (grid_fusion$seconds / grid_light$seconds - 1)))
cat(sprintf("Hierarchical, soft constraints vs light only: %+.1f%% wall clock\n",
            100 * (hier_fusion$seconds / hier_light$seconds - 1)))
cat(sprintf("FFBS, soft constraints vs light only        : %+.1f%% wall clock\n",
            100 * (ffbs_fusion$seconds / ffbs_light$seconds - 1)))
```

```{r masked-fraction}
# What fraction of the grid does the hard constraint actually exclude? That is
# the ceiling on the speed-up, because the saving is exactly the light
# likelihood not evaluated on excluded cells.
cells   <- terra::crds(grid)
sea_at  <- as.numeric(terra::extract(sea_rast, cells)[, 1])
dep_at  <- as.numeric(terra::extract(depth_rast, cells)[, 1])
deepest <- max(vapply(tags, function(g) max(g$light$depth_max, na.rm = TRUE), 0))
cat(sprintf("cells excluded by the land mask             : %.1f%%\n",
            100 * mean(sea_at == 0 | is.na(sea_at))))
cat(sprintf("cells too shallow for the deepest dive      : %.1f%% (deepest %.0f m)\n",
            100 * mean(dep_at < (deepest - params$bathy_margin_m), na.rm = TRUE), deepest))
cat(sprintf("union, i.e. ceiling on the achievable saving: %.1f%%\n",
            100 * mean(sea_at == 0 | is.na(sea_at) |
                       dep_at < (deepest - params$bathy_margin_m))))
```

```{r spread}
# The other half of the question: if fusion buys a tighter posterior for the same
# number of sweeps, it is cheaper at matched precision even when it is slower
# per sweep.
spread <- function(fit, label) data.frame(fit = label,
  mean_lon_sd = round(mean(vapply(fit$tracks, function(t) mean(t$lon_sd), 0)), 3),
  mean_lat_sd = round(mean(vapply(fit$tracks, function(t) mean(t$lat_sd), 0)), 3))
rbind(spread(hier_light$value, "hier_light"), spread(hier_fusion$value, "hier_fusion"))
```

# Population-level results

```{r population}
rbind(data.frame(fit = "light",     t(round(hier_light$value$population, 1))),
      data.frame(fit = "light+env", t(round(hier_fusion$value$population, 1))))
merge(hier_light$value$movement, hier_fusion$value$movement,
      by = "id", suffixes = c("_light", "_env"))
```

# Figures

```{r track-maps, fig.height = 11}
op <- par(mfrow = c(4, 3), mar = c(3.4, 3.4, 2.2, 0.6), mgp = c(2.1, 0.7, 0))
for (id in names(tags)) {
  g  <- tags[[id]]
  tl <- hier_light$value$tracks[[id]]
  tf <- hier_fusion$value$tracks[[id]]
  gf <- grid_fusion$value[[id]]
  xr <- range(c(lon360(tl$lon), lon360(tf$lon), g$argos$lon))
  yr <- range(c(tl$lat, tf$lat, g$argos$lat))
  plot(NA, xlim = xr, ylim = yr, xlab = "lon (E)", ylab = "lat",
       main = sprintf("%s / %s (%.0f d)", id, g$seal, g$days),
       asp = 1 / cos(mean(yr) * pi / 180))
  lines(g$argos$lon, g$argos$lat, col = "grey35", lwd = 2)
  lines(lon360(tl$lon), tl$lat, col = "#2980b9", lwd = 1.2)
  lines(lon360(tf$lon), tf$lat, col = "#c0392b", lwd = 1.2)
  lines(lon360(gf$lon), gf$lat, col = "#27ae60", lwd = 1.0, lty = 3)
  points(g$deploy[["lon"]], g$deploy[["lat"]], pch = 21, bg = "#2ecc71", cex = 1.2)
}
plot.new()
legend("center", bty = "n", cex = 0.95,
       legend = c("Argos (truth)", "hier, light", "hier, light+env",
                  "grid, light+env (hard)", "deployment"),
       col = c("grey35", "#2980b9", "#c0392b", "#27ae60", "#2ecc71"),
       lwd = c(2, 1.2, 1.2, 1, NA), lty = c(1, 1, 1, 3, NA),
       pch = c(NA, NA, NA, NA, 21), pt.bg = c(NA, NA, NA, NA, "#2ecc71"))
par(op)
```

```{r error-by-doy, fig.height = 5.5}
k <- knots[is.finite(knots$err_km) & knots$batch %in% c("light", "light+env"), ]
k$doy <- as.integer(format(k$time, "%j"))
brk <- seq(0, 378, by = 14); mid <- utils::head(brk, -1) + 7
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.4, 0.8))
for (eng in c("hier", "grid")) {
  ke <- k[k$engine == eng, ]
  if (!nrow(ke)) next
  plot(NA, xlim = range(ke$doy), ylim = c(0, stats::quantile(ke$err_km, .95)),
       xlab = "day of year", ylab = "great-circle error (km)", main = paste(eng, "engine"))
  rect(265 - 21, -1e4, 265 + 21, 1e4, col = "#f4ecf7", border = NA)   # September equinox
  for (b in c("light", "light+env")) {
    kb <- ke[ke$batch == b, ]; if (!nrow(kb)) next
    lines(mid, as.numeric(tapply(kb$err_km, cut(kb$doy, brk), stats::median)),
          col = if (b == "light") "#2980b9" else "#c0392b", lwd = 2)
  }
  legend("topleft", c("light", "light + env", "equinox +/- 21 d"),
         col = c("#2980b9", "#c0392b", "#f4ecf7"), lwd = c(2, 2, 8), bty = "n", cex = 0.9)
}
par(op)
```

```{r error-vs-depth, fig.height = 4.5}
# Does the error track how deep the animal was? This is the empirical form of
# "light at depth is less informative", and it is what motivates the bathymetry
# term doing double duty as an information source and as a pruning device.
kd <- knots[knots$engine == "hier" & knots$batch == "light" & is.finite(knots$err_km), ]
kd$maxdepth <- unlist(lapply(split(seq_len(nrow(kd)), kd$id), function(ix) {
  g <- tags[[kd$id[ix[1]]]]
  vapply(as.numeric(kd$time[ix]), function(tt) {
    sel <- g$light$time > tt - params$step_hours * 3600 & g$light$time <= tt
    if (any(sel)) max(g$light$depth_max[sel], na.rm = TRUE) else NA_real_
  }, 0)
}))
bins <- cut(kd$maxdepth, breaks = c(0, 200, 400, 600, 800, 1000, Inf))
op <- par(mar = c(4.4, 4.2, 2.4, 0.8))
boxplot(kd$err_km ~ bins, outline = FALSE, col = "#d6eaf8",
        xlab = "deepest dive in the knot (m)", ylab = "error (km)",
        main = "Position error vs diving depth (hierarchical, light only)")
par(op)
print(round(tapply(kd$err_km, bins, stats::median), 1))
```

# Reproducibility record

```{r provenance}
saveRDS(list(params = params, manifest = manifest, meta = meta, tag_table = tag_table,
             results = results, batch = batch, contrasts = contrasts, timing = timing,
             clock_table = clock_table, spot_drift = spot_drift,
             session = utils::sessionInfo()),
        file.path(params$out_dir, "analysis_record.rds"))
utils::sessionInfo()
```

## Notes for the manuscript

- Panel: 10 female post-moult northern elephant seals, Ano Nuevo 2021, 137-248
  day deployments, Mk9 archival light at 4 s thinned to 10 min, Argos ground
  truth. One instrument in the folder carried no light channel and was dropped.
- Median and RMSE error against Argos per engine and per batch (7.2), with the
  per-tag spread (7.1) and the latitude/longitude split.
- The paired sensor-fusion contrast (7.3), and specifically whether the gain is
  concentrated in latitude and in the equinox window, which is the mechanism
  Methods 2.5 claims.
- Wall clock per batch (section 8) with the excluded-cell fraction alongside.
  The claim to make is narrow and testable: a **hard** auxiliary constraint makes
  the grid HMM faster roughly in proportion to the fraction of candidate cells it
  excludes, because the light likelihood is never evaluated there; a soft
  constraint carrying the same information does not.
- Nominal 95% interval coverage (7.2, 7.3), the empirical counterpart to the
  simulation-based calibration reported on synthetic data.
- The clock-correction contrast (7.3) alongside the SPOT tags' measured drift
  (3.7), which shows what a clock of this vintage does over eight months.
- Still needed from the data provider: permission to archive, permit and ethics
  numbers, funders, and the authorship line.
